---
title: "Immune Subtypes"
author: "Qirui Zhang, Siming Zhao"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: hide
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

In total, **80 analyses** were conducted across **20 tumor types**, each classified into **4 distinct immune phenotype groups (C1–C4)**. Among these analyses, **70 completed successfully**, while **10 encountered errors**. The specific tumor type and phenotype combinations for unsuccessful analyses are detailed below.

# Summary of Unsuccessful Analyses

The table below summarizes analyses that failed, along with explanations for their failure:

```{r, echo=FALSE}
library(knitr)
library(kableExtra)
library(dplyr)

failure_data <- data.frame(
  Combination = c("C1_SKCM_mutations", "C2_SKCM_mutations", "C2_GBM_mutations", "C2_KICH_mutations",
                  "C3_GBM_mutations", "C3_CESC_mutations", "C3_LUSC_mutations", "C3_SKCM_mutations", 
                  "C3_UCS_mutations","C4_SKCM_mutations"),
  Reason = c("Insufficient data (only two samples)",
             "Insufficient data (only two samples)",
             rep("Lacked phenotype variation (all samples category 0)", 5),
             "Insufficient data (only two samples)",
             "Lacked phenotype variation (all samples category 0)",
             "Insufficient data (only two samples)")
)

kable(failure_data, format = "html", col.names = c("Phenotype and Tumor Type", "Reason for Failure"), align = 'l') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>%
  row_spec(0, bold = TRUE, background = "#337ab7", color = "white") 
```

# Extracting and Summarizing Analysis Results
This code segment scans through analysis output directories, identifies relevant result files (*_resdd.Rd), extracts corresponding tumor types, immune phenotypes, and analysis modes(sig/reg).
```{r}
get_files <- function(outputdir, tumors = NULL, phenotypes = NULL) {
  files <- list.files(
    path = outputdir,
    pattern = "_resdd\\.Rd$",
    recursive = TRUE,
    full.names = TRUE
  )
  
  matching_files_txt <- sub("\\.Rd$", ".txt", files)
  
  analysis_mode <- ifelse(grepl("_sig_", basename(files)), "sig", "reg")

  inner_folder_names      <- basename(dirname(files))            
  phenotype_folder_names  <- basename(dirname(dirname(files)))  
  
  tumor_names <- mapply(function(inner_name, phen) {
    pattern_front <- paste0("^", phen, "_")
    tmp <- sub(pattern_front, "", inner_name)
    tmp <- sub("_mutations$", "", tmp)
    return(tmp)
  }, inner_folder_names, phenotype_folder_names)
  
  phenos_all <- phenotype_folder_names
  
  filedf <- data.frame(
    tumor      = tumor_names,
    phenotype  = phenos_all,
    mode       = analysis_mode,   
    file       = files,
    filetxt    = matching_files_txt,
    stringsAsFactors = FALSE
  )
  
  if (!is.null(tumors)) {
    filedf <- filedf[filedf$tumor %in% tumors, ]
  }
  if (!is.null(phenotypes)) {
    filedf <- filedf[filedf$phenotype %in% phenotypes, ]
  }

  rownames(filedf) <- NULL
  return(filedf)
}

filedf <- get_files(
  outputdir = "/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/temp/output/new_plot/immunesubtypes"
)
```

# Summary Table of Significant Differential Genes
The provided function extracts and compiles results from analyses, filtering specifically for significant differential genes based on a threshold (**`dd.fdr < 0.1`**).  
```{r}
get_diff_table <- function(filedf, include_nonsig = TRUE){
  pheno_all <- unique(filedf$phenotype)
  numlist <- list()
  
  for (p in pheno_all){
    p_txtfiles <- filedf[filedf$phenotype == p, ]
    numlist[[p]] <- list()
    
    for (t in seq_len(nrow(p_txtfiles))){
      txtf  <- p_txtfiles[t, "filetxt"]
      rdf   <- p_txtfiles[t, "file"] 
      tumor <- p_txtfiles[t, "tumor"] 
      mode  <- p_txtfiles[t, "mode"]     # reg / sig
      
      env <- new.env()
      load(rdf, envir = env)
      res_rdata <- env$res
      
      res <- read.table(txtf, header = TRUE)
      res$gene <- row.names(res)
      res$mode <- mode
      
      res$alpha <- sapply(res$gene, function(gene){
          res_rdata[[gene]][["dd"]][["res.alt"]]$alpha[2]
      })
      
      res$significant <- res$dd.fdr < 0.1
      
      if (include_nonsig) {
        numlist[[p]][[ paste0(tumor, "_", mode) ]] <- res
      } else {
        sig_res <- res[res$dd.fdr < 0.1, ]
        if (nrow(sig_res) > 0){
          numlist[[p]][[ paste0(tumor, "_", mode) ]] <- sig_res
        }
      }
    }
    
    if (length(numlist[[p]]) == 0){
      numlist[[p]] <- NULL
    }
  }
  
  combined_df <- do.call(
    rbind,
    lapply(names(numlist), function(pheno) {
      numlist.pheno <- numlist[[pheno]]
      df.pheno <- do.call(rbind, lapply(names(numlist.pheno), function(tname) {
        df <- numlist.pheno[[tname]]
        df$tumor <- sub("_.*", "", tname) 
        df$mode  <- sub(".*_", "", tname) 
        return(df)
      }))
      df.pheno$pheno <- pheno
      return(df.pheno)
    })
  )
  return(combined_df)
}

diff_table_all <- get_diff_table(filedf, include_nonsig = TRUE)
diff_table <-  get_diff_table(filedf, include_nonsig = FALSE)
```

```{r}
library(DT)
diff_table_sig <- subset(diff_table, mode == "sig")
datatable(
  diff_table_sig,
  extensions = 'Buttons',
  options = list(
    dom        = 'Bfrtip', 
    buttons    = c('copy', 'csv', 'excel', 'pdf', 'print'),
    pageLength = 10
  )
)
```

