---
title: "Simulation Power Results"
author: "Jie Zhou, Siming Zhao"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: hide
---

## Overview

For 200 genes with five functional annotations, we compare the power (sensitivity)
and false positive rate (FPR) of DiffDriver against benchmark methods: Linear
regression, Fisher's exact test, Binomial test, and Logistic Regression.

The first 50 genes have differential selection while the remaining 150 do not.
FDR is controlled at 0.1 using the BH procedure.

### Simulation parameters

These simulations use the same pipeline described in
[Simulation Code](simulation_code.html). The key parameters are:

| Parameter | Description |
|-----------|-------------|
| `betaf0`  | Baseline functional effect size: 0 or 1 |
| `para`    | Binary: selection probabilities c(prob_case, prob_control). Continuous: c(mean, sd, intercept, slope) |
| `beta_gc` | Functional effect coefficients (TSGpars or OGpars) |
| `Nsample` | Varies: 200, 400, 600, 800, 1000, 1200, 1500 |
| `Niter`   | 100 (binary) or 50 (continuous) per gene |

### Functional effect coefficients

- **TSG parameters**: `(Intercept)=0.851, functypecode8=0.458, mycons=0.137, sift=0.124, phylop100=0.083, MA=0.306`
- **OG parameters**: `(Intercept)=0.450, functypecode8=-0.186, mycons=0.316, sift=0.020, phylop100=0.256, MA=0.140`

---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

METHODS      <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
MYCOL        <- c("#FF3D2E", "#00AFBB", "#aa00ff", "#0001f6", "#fdae61")
FDR          <- 0.1
BINARY_SIZES <- c(200, 400, 600, 800, 1000, 1200, 1500)
CONTI_SIZES  <- c(200, 400, 600, 800, 1000, 1200)

#' Compute power and FPR from p-value matrices across simulation iterations.
#' @param m.pvalue List of matrices (one per method), each n_genes x Nsim.
#' @param Nsim Number of simulation iterations.
#' @param n_true Number of true positive genes (first n_true genes).
compute_metrics <- function(m.pvalue, Nsim, n_true = 50) {
  n_methods <- length(m.pvalue)
  power <- fpr <- numeric(n_methods)
  for (k in seq_len(n_methods)) {
    sens <- fpr_vals <- numeric(Nsim)
    for (j in seq_len(Nsim)) {
      hits <- which(p.adjust(m.pvalue[[k]][, j], method = "BH") < FDR)
      if (length(hits) > 0) {
        sens[j]     <- sum(hits <= n_true) / n_true
        fpr_vals[j] <- sum(hits > n_true) / length(hits)
      }
    }
    power[k] <- mean(sens)
    fpr[k]   <- mean(fpr_vals)
  }
  list(power = power, fpr = fpr)
}

#' Load pre-computed simulation results and compute power/FPR per sample size.
#'
#' Each sample size x gene combination has two result files in subdirectories
#' diffFix/ and other/ under base_path. The p-values are extracted as:
#'   DiffDriver = simuresdiffFix$m1.pvalue
#'   Linear/Fisher/Binomial/LogisticR = simuresother$m1--m4.pvalue
#'
#' @param base_path Directory containing diffFix/ and other/ subdirectories.
#' @param file_fn Function(Nsample, gene_index) returning the .Rd filename.
#' @param sample_sizes Vector of sample sizes to evaluate.
#' @param Nsim Max number of iterations to use. If NULL, uses all available.
#' @param n_genes Number of genes (default 200).
load_and_compute <- function(base_path, file_fn, sample_sizes,
                             Nsim = NULL, n_genes = 200) {
  n_methods <- length(METHODS)
  power_mat <- fpr_mat <- matrix(
    nrow = n_methods, ncol = length(sample_sizes),
    dimnames = list(METHODS, sample_sizes)
  )
  for (s in seq_along(sample_sizes)) {
    N <- sample_sizes[s]
    m.pvalue <- NULL
    n_iter   <- NULL
    for (g in seq_len(n_genes)) {
      fname <- file_fn(N, g)
      load(file.path(base_path, "diffFix", fname))
      load(file.path(base_path, "other", fname))
      if (is.null(n_iter)) {
        n_avail <- length(simuresdiffFix[["m1.pvalue"]])
        n_iter  <- if (!is.null(Nsim)) min(Nsim, n_avail) else n_avail
        m.pvalue <- lapply(seq_len(n_methods), function(x)
          matrix(nrow = n_genes, ncol = n_iter))
      }
      idx <- seq_len(n_iter)
      m.pvalue[[1]][g, ] <- simuresdiffFix[["m1.pvalue"]][idx]
      m.pvalue[[2]][g, ] <- simuresother[["m1.pvalue"]][idx]
      m.pvalue[[3]][g, ] <- simuresother[["m2.pvalue"]][idx]
      m.pvalue[[4]][g, ] <- simuresother[["m3.pvalue"]][idx]
      m.pvalue[[5]][g, ] <- simuresother[["m4.pvalue"]][idx]
    }
    metrics <- compute_metrics(m.pvalue, n_iter)
    power_mat[, s] <- metrics$power
    fpr_mat[, s]   <- metrics$fpr
  }
  list(power = power_mat, fpr = fpr_mat)
}

#' Plot power and FPR side by side.
#' @param res Output from load_and_compute (list with power and fpr matrices).
#' @param power_title Title for the power panel.
#' @param fpr_title Title for the FPR panel.
#' @param power_ylim Y-axis upper limit for the power panel.
#' @param show_legend Whether to show the method legend on the power panel.
plot_power_fpr <- function(res, power_title, fpr_title,
                           power_ylim = 0.4, show_legend = TRUE) {
  x     <- seq_len(ncol(res$power))
  sizes <- colnames(res$power)
  par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))

  # Power panel
  plot(x, ylim = c(0, power_ylim), col = "white", ylab = "Power",
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = "n",
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title(power_title)
  yticks <- seq(0, power_ylim, 0.1)
  axis(2, at = yticks, lab = paste0(yticks * 100, "%"), las = 2, cex.axis = 1.3)
  axis(1, at = x, lab = sizes, cex.axis = 1.3)
  grid()
  for (r in seq_len(nrow(res$power))) {
    points(x, res$power[r, ], col = MYCOL[r], pch = 19, cex = 1.8)
    lines(x, res$power[r, ], col = MYCOL[r], lwd = 1.5)
  }
  if (show_legend) {
    legend("topleft", legend = METHODS, col = MYCOL[seq_along(METHODS)],
           lwd = 1, bty = "n", lty = NA, pch = 19)
  }

  # FPR panel
  plot(x, ylim = c(0, 1), col = "white", ylab = "FPR",
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = "n",
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title(fpr_title)
  axis(2, at = seq(0, 1, 0.1), lab = paste0(seq(0, 100, 10), "%"),
       las = 2, cex.axis = 1.3)
  axis(1, at = x, lab = sizes, cex.axis = 1.3)
  grid()
  for (r in seq_len(nrow(res$fpr))) {
    points(x, res$fpr[r, ], col = MYCOL[r], pch = 19, cex = 1.8)
    lines(x, res$fpr[r, ], col = MYCOL[r], lwd = 1.5)
  }
}
```

## TSG, Binary Phenotype

### para = c(0.8, 0.2)

```{r tsg-binary-08, fig.width=12, fig.height=5}
res <- load_and_compute(
  base_path    = "/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA",
  file_fn      = function(N, g) paste0("pppower_betaf0=0_sample", N, "_", g, ".Rd"),
  sample_sizes = BINARY_SIZES
)
plot_power_fpr(res,
  power_title = "para = (0.8, 0.2)",
  fpr_title   = "para = (0.5, 0.5), FDR = 0.1"
)
```

### para = c(0.9, 0.1)

```{r tsg-binary-09, fig.width=12, fig.height=5}
res <- load_and_compute(
  base_path    = "/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/TSG0.9/TSG_gene",
  file_fn      = function(N, g) paste0("pppower_betaf0=0_sample", N, "_", g, ".Rd"),
  sample_sizes = BINARY_SIZES
)
plot_power_fpr(res,
  power_title  = "para = (0.9, 0.1)",
  fpr_title    = "para = (0.5, 0.5), FDR = 0.1",
  show_legend  = FALSE
)
```

---

## TSG, Continuous Phenotype

Continuous phenotype with `para = c(0, 2, 0, 1)`: phenotype ~ N(0, 2), selection
probability determined by logistic regression with intercept = 0 and slope = 1.

### betaf0 = 0

```{r tsg-conti-f0, fig.width=12, fig.height=5}
res <- load_and_compute(
  base_path    = "/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2",
  file_fn      = function(N, g) paste0("pppower_betaf0=0_sample", N, "_", g, ".Rd"),
  sample_sizes = CONTI_SIZES
)
plot_power_fpr(res,
  power_title = "betaf0 = 0, para = (0, 2, 0, 1)",
  fpr_title   = "para = (0, 2, 0, 0), FDR = 0.1",
  power_ylim  = 0.6
)
```

### betaf0 = 1

```{r tsg-conti-f1, fig.width=12, fig.height=5}
res <- load_and_compute(
  base_path    = "/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2",
  file_fn      = function(N, g) paste0("pppower_betaf0=1_sample", N, "_", g, ".Rd"),
  sample_sizes = CONTI_SIZES
)
plot_power_fpr(res,
  power_title  = "betaf0 = 1, para = (0, 2, 0, 2)",
  fpr_title    = "para = (0, 2, 0, 0), FDR = 0.1",
  power_ylim   = 0.6,
  show_legend  = FALSE
)
```

---

## OG, Binary Phenotype

### para = c(0.8, 0.2)

```{r og-binary-08, fig.width=12, fig.height=5}
res <- load_and_compute(
  base_path    = "/Volumes/Szhao/jiezhou/diffdata/binaryEM/OG_gene",
  file_fn      = function(N, g) paste0("pppower_betaf0=0_sample", N, "_0.8_", g, ".Rd"),
  sample_sizes = BINARY_SIZES
)
plot_power_fpr(res,
  power_title = "para = (0.8, 0.2)",
  fpr_title   = "para = (0.5, 0.5), FDR = 0.1",
  power_ylim  = 0.5
)
```

### para = c(0.9, 0.1)

```{r og-binary-09, fig.width=12, fig.height=5}
res <- load_and_compute(
  base_path    = "/Volumes/Szhao/jiezhou/diffdata/binaryEM/OG_gene",
  file_fn      = function(N, g) paste0("pppower_betaf0=0_sample", N, "_0.9_", g, ".Rd"),
  sample_sizes = BINARY_SIZES
)
plot_power_fpr(res,
  power_title  = "para = (0.9, 0.1)",
  fpr_title    = "para = (0.5, 0.5), FDR = 0.1",
  power_ylim   = 0.5,
  show_legend  = FALSE
)
```

---

## OG, Continuous Phenotype

Continuous phenotype with `para = c(0, 2, 0, 1)`: phenotype ~ N(0, 2), selection
probability determined by logistic regression with intercept = 0 and slope = 1.

### betaf0 = 0

```{r og-conti-f0, fig.width=12, fig.height=5}
res <- load_and_compute(
  base_path    = "/Volumes/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene",
  file_fn      = function(N, g) paste0("pppower_betaf0=0_sample", N, "_", g, ".Rd"),
  sample_sizes = CONTI_SIZES
)
plot_power_fpr(res,
  power_title = "betaf0 = 0, para = (0, 2, 0, 1)",
  fpr_title   = "para = (0, 2, 0, 0), FDR = 0.1",
  power_ylim  = 0.6
)
```

### betaf0 = 1

```{r og-conti-f1, fig.width=12, fig.height=5}
res <- load_and_compute(
  base_path    = "/Volumes/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene",
  file_fn      = function(N, g) paste0("pppower_betaf0=1_sample", N, "_", g, ".Rd"),
  sample_sizes = CONTI_SIZES
)
plot_power_fpr(res,
  power_title  = "betaf0 = 1, para = (0, 2, 0, 2)",
  fpr_title    = "para = (0, 2, 0, 0), FDR = 0.1",
  power_ylim   = 0.6,
  show_legend  = FALSE
)
```
