---
title: "Simulation: Estimated Background under Signature Confounding"
author: "Jie Zhou, Siming Zhao"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: hide
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This is a stricter version of the
[phenotype–signature confounding simulation](Simulation-BMR.html). The setup is
identical — the phenotype correlates with a mutational signature (LUAD signature
3, $R^2 = 0.78$) but has **zero** correlation with selection, and mutations are
simulated for a single gene (ERBB3) with hotspots at the signature's peak context
(nucleotide type 34) — so any detection is a false positive.

The difference is how the two background-modeling methods obtain the background
mutation rate:

- In [Simulation-BMR](Simulation-BMR.html) both DiffDriver and coselens are
  **given the true** per-context background rate.
- **Here they must *estimate* it** from a simulated genome-wide **silent
  (synonymous) mutation count matrix**:
    - **DiffDriver** estimates its background by running its signature model
      (`fastTopics` on the 96 × N count matrix, exactly as
      `diffdriver::matrixlistToBMR` does) to recover the per-sample per-context
      spectrum.
    - **coselens** estimates its per-context substitution rate as
      counts / sites — `N_syn / Lsyn96` per group — which is exactly the
      Poisson fit `dndscv` performs; this rate is fed to `dndscv_truerate`.

The question is whether estimating the background from a finite, confounded
silent spectrum still controls the false positive rate the way the true rate did.

### Realistic LUAD background rate

The silent count matrix is not arbitrary — it is simulated from a realistic
LUAD background:

$$Y[c,s] \sim \text{Poisson}\big(\text{Lsyn96}[c]\cdot \text{bmr}[c,s]\cdot g\big),$$

where

- **`bmr = (signatures %*% loadings) / sc`** uses the **real LUAD** mutational
  signatures (96 × 6) and per-sample loadings (`ll$LUAD`), so the spectrum shape
  is exactly LUAD's; the mean per-site probability is ≈ 1.8 × 10⁻⁶.
- **`Lsyn96`** is the number of synonymous sites per trinucleotide context
  (from dndscv's RefCDS; `data/Lsyn.rds`), ≈ 23.3 M sites genome-wide.
- **`g`** calibrates the mean per-sample synonymous burden to the **observed
  LUAD burden of ≈ 64 synonymous mutations per sample** (530 TCGA-LUAD samples,
  from `LUAD_mutations.txt` annotated with dndscv).

Crucially, the true `bmr` (with `sc = 359000`) already produces ≈ 63 synonymous
mutations per sample, so **`g = 64.2 / 63 ≈ 1.02`** — essentially 1. The silent
count matrix and the ERBB3 mutations therefore share **one realistic LUAD-scale
background**, with no rescaling needed.

Both methods were run on the **same simulated cohorts** as the true-background
simulation (the cohort RNG stream is frozen around the estimation step, so the
per-iteration mutation counts are identical).

## Rerunning

The estimated-background helpers are in
[`code/simulate_silent_estimate.R`](https://github.com/szhaolab/diffdriver-paper/blob/main/code/simulate_silent_estimate.R):
`build_silent_matrix()` (the LUAD-calibrated silent count matrix),
`coselens_estimated_rate()` (`N_syn/Lsyn96` per group → `dndscv_truerate`), and
`diffdriver_estimated_mr()` (fastTopics spectrum → per-site log-background). Per
iteration:

```{r eval=FALSE}
source("code/simulate_functions.R"); source("code/simulate_silent_estimate.R")
source("code/run_coselens.R"); source("code/dndscv_truerate.R")
library(diffdriver); library(coselens)
load("data/bmr_sim_inputs.rda"); load("data/hmm.rda")
load("data/codeSignature.rda");  load("data/hotspot1sig.rda")
cm <- readRDS("data/context96_map.rds"); Lsyn96 <- readRDS("data/Lsyn.rds")$Lsyn96

sim <- simulate_confounding(binary = TRUE, sganno = bmr_sim$sganno,
  sgmatrix = bmr_sim$sgmatrix, Nsample = 1500, para = c(0.5, 0.5),
  beta_gc = bmr_sim$beta_gc, signatures = bmr_sim$signature, loadings = bmr_sim$ll,
  rho = sqrt(0.78), hot = 1, hmm = hmm, sc = 359000)
bmr <- sim$bmrfold$bmr; e <- sim$pheno; g1 <- which(e == 1); g2 <- which(e == 0)
maf <- sim_to_maf(sim$mutlist, sim$annodata, e)

Y <- build_silent_matrix(bmr, Lsyn96, target_silent = 64.2)   # ~64 syn/sample

# coselens: estimated per-group rate from the silent counts
mr1 <- coselens_estimated_rate(Y, g1, Lsyn96, cm)
mr2 <- coselens_estimated_rate(Y, g2, Lsyn96, cm)
cos <- run_coselens(maf$group1, maf$group2, subset.genes.by = "ERBB3",
         true_rate = list(mutrates1 = mr1, mutrates2 = mr2, theta = 1e6),
         gene_list = "ERBB3", cv = NULL)$substitutions$pval

# DiffDriver: estimated per-site background from fastTopics(Y)
mr_est <- do.call(rbind, diffdriver_estimated_mr(Y, sim$bmrmtxlist, bmr))
dd <- ddmodel_binary(do.call(rbind, sim$mutlist), e, mr_est, sim$efsize$diffFe)$pvalue
```

The full benchmark loops this over `Nsample = 200 … 1500` (100 iterations each)
and stores the results in `data/pppower_estimatedBMR_bmr.Rd`.

## Results

Both DiffDriver and coselens **control the false positive rate whether they are
given the true background or estimate it** from the LUAD-calibrated silent count
matrix (all curves stay near zero). The count-based benchmarks (Linear, Fisher,
Binomial, Logistic regression, Mann-Whitney) do **not** use or estimate a
background rate — they operate on the raw mutation counts, so the "true vs
estimated" distinction does not apply to them. They are shown only for context
(taken from the [true-background simulation](Simulation-BMR.html), on the same
setup) and rise toward ~0.6 with sample size.

```{r fig.width=8, fig.height=6}
Nsim <- 100
SIZES <- c(200,400,600,800,1000,1200,1500)
load("data/pppower_estimatedBMR_bmr.Rd")     # object `estBMR`, keyed by N
load("data/pppower_mannwhitney_bmr.Rd")      # object `mw`

fp <- function(p) mean(p < 0.01, na.rm = TRUE)
rate <- function(field) sapply(as.character(SIZES), function(n) fp(estBMR[[n]][[field]]))

# background-modeling methods (this simulation): 4 series
cosT <- rate("coselens_true"); cosE <- rate("coselens_est")
ddT  <- rate("diff_true");     ddE  <- rate("diff_est")

# count-based benchmarks (for context; from the true-background sim, same setup)
lin <- fis <- bin <- lr <- mw_fp <- numeric(length(SIZES))
for (i in seq_along(SIZES)) {
  N <- SIZES[i]
  o <- new.env(); load(sprintf("data/bmr_archive/other/pppower_betaf0=1_sample%d_0.5_0.Rd", N), envir = o)
  s <- o$simuresother
  lin[i] <- fp(s$m1.pvalue); fis[i] <- fp(s$m2.pvalue); bin[i] <- fp(s$m3.pvalue); lr[i] <- fp(s$m4.pvalue)
  mw_fp[i] <- fp(mw[[as.character(N)]]$mannwhitney.pvalue)
}

x <- seq_along(SIZES)
par(mar = c(5,5,3,9), xpd = NA)
plot(x, ylim = c(0,0.7), col = "white", ylab = "False positive rate (p < 0.01)",
     xlab = "Number of samples", xaxt = "n", las = 1)
axis(1, at = x, labels = SIZES); grid(); abline(h = 0.01, lty = 3, col = "grey60")

# count-based (grey-ish, thin) for context
cb <- list(Linear = lin, Fisher = fis, Binomial = bin, LogisticR = lr, MannWhitney = mw_fp)
cbcol <- c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#1a9850")
for (j in seq_along(cb)) { lines(x, cb[[j]], col = cbcol[j], lwd = 1.2); points(x, cb[[j]], col = cbcol[j], pch = 20, cex = 0.9) }

# background-modeling: coselens (orange) / DiffDriver (red); true = filled/solid, est = open/dashed
lines(x, cosT, col = "#ff7f00", lwd = 2);              points(x, cosT, col = "#ff7f00", pch = 19, cex = 1.4)
lines(x, cosE, col = "#ff7f00", lwd = 2, lty = 2);     points(x, cosE, col = "#ff7f00", pch = 1,  cex = 1.4, lwd = 2)
lines(x, ddT,  col = "#FF3D2E", lwd = 2);              points(x, ddT,  col = "#FF3D2E", pch = 17, cex = 1.4)
lines(x, ddE,  col = "#FF3D2E", lwd = 2, lty = 2);     points(x, ddE,  col = "#FF3D2E", pch = 2,  cex = 1.4, lwd = 2)

legend(par("usr")[2], par("usr")[4], bty = "n", cex = 0.85, xjust = 0,
  legend = c("Coselens (true bkg)","Coselens (estimated)",
             "DiffDriver (true bkg)","DiffDriver (estimated)","",
             "Linear","Fisher","Binomial","LogisticR","MannWhitney"),
  col = c("#ff7f00","#ff7f00","#FF3D2E","#FF3D2E",NA, cbcol),
  pch = c(19,1,17,2,NA, rep(20,5)),
  lty = c(1,2,1,2,NA, rep(1,5)), lwd = c(2,2,2,2,NA, rep(1.2,5)))
title("Estimated vs. true background under signature confounding")
```

The two dashed curves (estimated background) sit right on top of their solid
counterparts (true background), near zero — estimating the background from a
realistic, LUAD-calibrated silent count matrix does not reintroduce the
confounding, for either DiffDriver or coselens.
