---
title: "Simulation: Confounding Between Phenotype and Mutational Signatures"
author: "Jie Zhou, Siming Zhao"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: hide
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This simulation demonstrates that when the phenotype is correlated with
mutational signatures — but **not** with selection — benchmark methods produce
false positives while DiffDriver correctly controls the false positive rate.

### Simulation setup

- **Background mutation rate (realistic, LUAD-derived)**: The per-site
  background probability is `bmr = (signatures %*% loadings) / sc`, built entirely
  from LUAD data — so it is *not* an arbitrary rate:
    - **Shape**: the `signatures` (96 × 6) and per-sample `loadings` (`ll$LUAD`)
      are the real LUAD mutational signatures and exposures (fastTopics on the
      LUAD mutation counts, 527 samples, bootstrapped up to 1,500). The loadings
      are a composition (constant column sums = 62.75), so a summed-loading
      covariate carries no phenotype information and cannot adjust the confounding
      away.
    - **Level**: `sc = 359000` gives a mean per-site probability of ≈1.8×10⁻⁶ and
      a genome-wide **synonymous burden of ≈63 mutations per sample** (median 62,
      range 56–73), matching the **observed ≈64 synonymous mutations per sample**
      in real TCGA-LUAD (530 samples). So both the spectrum and the overall level
      are LUAD-realistic.
- **Mutational signatures**: Six signatures extracted from LUAD using
  [fastTopics](https://stephenslab.github.io/fastTopics/). Signature 3 — a
  T>G–dominated (smoking-like) signature whose highest-loading context is
  `C[T>G]G` (nucleotide type 34) — is selected as the confounding signature.
- **Phenotype**: Simulated to correlate with signature 3 ($R^2 = 0.78$), but
  with **zero correlation** with selection (`para = c(0.5, 0.5)`).
- **Mutations**: Simulated for a single gene (ERBB3) using oncogene mutation
  rates (background + selection), with two hotspots placed at nucleotide type 34
  (the context with the highest loading in signature 3), so that the
  hotspot-dominated mutations track the phenotype.

### Key result

Because the phenotype correlates with a mutational signature that affects
overall mutation rates, the count-based benchmark methods (Linear regression,
Fisher's exact test, Binomial test, Logistic regression, and the Mann-Whitney
rank-sum test) increasingly detect differential selection as sample size grows —
all false positives. DiffDriver, which accounts for signature-based background
mutation rates, maintains a low false positive rate.

[coselens](https://github.com/ggonzalezlopez/coselens) is also included as a
"true-background" reference: the phenotype is binarized into two groups and
coselens is given each group's **real** mutation rate (the group-summed neutral
rate per trinucleotide context from the simulation, assuming the rate is the same
across samples within a group; see [coselens setup](#coselens-with-the-true-per-group-rate)).
Because it is handed the true confounded background, coselens correctly
attributes the case/control mutation excess to background rather than selection
and also controls the false positive rate.

### Robustness across realistic variations

The [Signature-Confounding under Realistic Variations](Simulation-BMR-variations.html)
page stress-tests this benchmark under variable per-sample mutation burden,
different confounding strengths ($R^2$), a continuous phenotype, and DiffDriver's
robustness to the number of fitted signatures. The count-based benchmarks stay
vulnerable throughout, while DiffDriver and Coselens keep the false positive rate
controlled.

## Rerunning the simulation

The simulation functions live in [`code/simulate_functions.R`](https://github.com/szhaolab/diffdriver-paper/blob/main/code/simulate_functions.R)
under the "CONFOUNDING scenario" section (`build_confounded_bmr`,
`simulate_confounding`, `run_confounding_benchmarks`,
`run_confounding_diffdriver`, `run_confounding_diffdriver_fixed`). The benchmark
methods are all in `run_confounding_benchmarks`
(Linear/Fisher/Binomial/LogisticR + Mann-Whitney); DiffDriver is run by
`run_confounding_diffdriver_fixed`.

The inputs are bundled in `data/bmr_sim_inputs.rda` (`beta_gc`, `beta_gcFix`,
`signature`, `ll`, `sganno`, `sgmatrix`), together with `data/hmm.rda`,
`data/codeSignature.rda`, and `data/hotspot1sig.rda` (hotspots at nttype 34).
A single sample size is reproduced with:

```{r eval=FALSE}
library(diffdriver); library(data.table); library(Matrix)
source("code/simulate_functions.R")
load("data/bmr_sim_inputs.rda"); load("data/hmm.rda")
load("data/codeSignature.rda");  load("data/hotspot1sig.rda")

set.seed(1)
res_other <- run_confounding_benchmarks(
  binary = TRUE, Niter = 100,
  sganno = bmr_sim$sganno, sgmatrix = bmr_sim$sgmatrix,
  Nsample = 1500, para = c(0.5, 0.5),      # no differential selection
  rho = sqrt(0.78),                        # phenotype ~ signature 3, R^2 = 0.78
  signatures = bmr_sim$signature, loadings = bmr_sim$ll,
  beta_gc = bmr_sim$beta_gc, hot = 1, hmm = hmm,
  sc = 359000)                             # background-rate scale (~188 mutations)

# false-positive rate per method (p < 0.01), no true signal so all are FPs
sapply(res_other[c("m1.pvalue","m2.pvalue","m3.pvalue","m4.pvalue",
                    "mannwhitney.pvalue")], function(p) mean(p < 0.01))
```

The full benchmark loops this over `Nsample = 200 … 1500`; the Mann-Whitney
p-values used below were produced this way and saved to
`data/pppower_mannwhitney_bmr.Rd`. Because the original
`scripts/simulate_functions.R` (and the exact HMM coefficients) were lost, the
rerun is *close* to — but not bit-identical to — the archived runs; the Linear
and Fisher/Binomial/LogisticR curves plotted below are read from the original
archived results.

### coselens with the true per-group rate

coselens compares dN/dS between two groups. We binarize the phenotype into cases
(E = 1) and controls (E = 0) and, rather than letting dndscv estimate the
background, feed each group its **true** neutral rate from the simulation. Under
the assumption that the rate is the same across samples within a group, a group's
per-context rate is the group-summed neutral rate `rowSums(bmrfold$bmr[, group])`;
this 96-context vector is mapped to dndscv's 192 substitution contexts via
`data/context96_map.rds` and passed to `dndscv_truerate` (a rate-fixed dndscv).
`code/run_coselens.R` wraps this; the per-iteration driver is:

```{r eval=FALSE}
source("code/run_coselens.R"); source("code/dndscv_truerate.R")
library(coselens)
context96_map <- readRDS("data/context96_map.rds")

# `sim` is one cohort returned by simulate_confounding() (same set.seed(1) loop
# as above, so the cohorts are identical to the other methods).
e   <- sim$pheno; g1 <- which(e == 1); g2 <- which(e == 0)
maf <- sim_to_maf(sim$mutlist, sim$annodata, e)           # case/control MAFs
mr1 <- rowSums(sim$bmrfold$bmr[, g1])[context96_map$number]  # 192-context true rate
mr2 <- rowSums(sim$bmrfold$bmr[, g2])[context96_map$number]
res <- run_coselens(maf$group1, maf$group2, subset.genes.by = "ERBB3",
                    true_rate = list(mutrates1 = mr1, mutrates2 = mr2, theta = 1e6),
                    gene_list = "ERBB3", cv = NULL)
res$substitutions$pval          # differential-selection p-value for ERBB3
```

coselens was run on the **same simulated cohorts** as the other methods (verified:
per-iteration mutation counts are identical across all 100 iterations and 7 sample
sizes). Its p-values are saved to `data/pppower_coselens_bmr.Rd`.



```{r}
Nsim=100
SIZES <- c(200,400,600,800,1000,1200,1500)

# Mann-Whitney and coselens FPs from the rerun on the SAME simulated cohorts
# (code/simulate_functions.R::run_confounding_benchmarks + code/run_coselens.R).
load("data/pppower_mannwhitney_bmr.Rd")   # object `mw`,       keyed by sample size
load("data/pppower_coselens_bmr.Rd")      # object `coselens`, keyed by sample size

mylist <- list()
for (i1 in c(1)){
   for (rho in c(0)) {
    for (i4 in c(0.5)) {
    n<-0
    for(i3 in SIZES){
      n <- n+1
      load(paste0("data/bmr_archive/diff/pppower_betaf0=",i1, "_sample",i3,"_",i4,"_",rho,".Rd"))
          load(paste0("data/bmr_archive/other/pppower_betaf0=",i1,"_sample",i3,"_",i4,"_",rho,".Rd"))

      m5.pvalue <- simuresdiff[["m1.pvalue"]]
      m1.pvalue <- simuresother[["m1.pvalue"]]
      m2.pvalue <- simuresother[["m2.pvalue"]]
      m3.pvalue <- simuresother[["m3.pvalue"]]
      m4.pvalue <- simuresother[["m4.pvalue"]]
      m7.pvalue <- mw[[as.character(i3)]][["mannwhitney.pvalue"]]       # Mann-Whitney
      m8.pvalue <- coselens[[as.character(i3)]][["coselens.pvalue"]]    # coselens (true per-group BMR)
     mylist[[n]] <- c(length(m1.pvalue[m1.pvalue <0.01]),length(m2.pvalue[m2.pvalue <0.01]),length(m3.pvalue[m3.pvalue <0.01]), length(m4.pvalue[m4.pvalue <0.01]),length(m7.pvalue[m7.pvalue <0.01]),length(m8.pvalue[m8.pvalue <0.01]),length(m5.pvalue[m5.pvalue <0.01]))/Nsim
    }
    outdt <- do.call(cbind, mylist)
    rownames(outdt)=c("Linear","Fisher","Binomial","LogisticR","MannWhitney","Coselens","diffDriver")
    colnames(outdt) <- SIZES
    # colors: Linear, Fisher, Binomial, LogisticR, MannWhitney, coselens, diff
    mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#1a9850","#ff7f00","#FF3D2E")
    mypch=c(19,19,19,19,3,8,17)
     plot(1:dim(outdt)[2],ylim=c(0,1), col="white", ylab = "False positive rate (p < 0.01)", xlab="Number of samples", yaxt="n", xaxt="n")
     title( paste( "SC(scale efficient)=",i1,",","para=(",i4,1-i4,")", ",", sep=" "))
    yticks_val <- seq(0,1,0.1)
    axis(2, at=yticks_val, lab=paste0(yticks_val),las=2)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt))
    grid()
    for (r in 1:7){
      points(outdt[r,],col=mycol[r], pch=mypch[r], cex=1.3)
      lines(outdt[r,],col=mycol[r])
    }
    legend("topleft", legend = c("Linear","Fisher","Binomial","LogisticR","MannWhitney","Coselens","diff"), col= mycol, lwd=1, bty="n", lty=rep(NA,7), pch=mypch)
  }
}
  }

```

## Estimated background: a stricter test

The result above hands DiffDriver and Coselens the **true** per-context background
rate. A stricter test is to make them **estimate** the background from a simulated
genome-wide **silent (synonymous) mutation count matrix**, rather than being given
it:

- **DiffDriver** estimates its background by running its signature model
  (`fastTopics` on the 96 × N silent count matrix, exactly as
  `diffdriver::matrixlistToBMR` does) to recover the per-sample per-context
  spectrum.
- **Coselens** estimates its per-context substitution rate as counts / sites —
  `N_syn / Lsyn96` per group — which is exactly the Poisson fit `dndscv` performs;
  this rate is fed to `dndscv_truerate`.

The question is whether estimating the background from a finite, confounded silent
spectrum still controls the false positive rate the way the true rate did.

### Realistic LUAD silent count matrix

The silent count matrix is not arbitrary — it is simulated from the same realistic
LUAD background:

$$Y[c,s] \sim \text{Poisson}\big(\text{Lsyn96}[c]\cdot \text{bmr}[c,s]\cdot g\big),$$

where `bmr = (signatures %*% loadings) / sc` uses the **real LUAD** signatures
(96 × 6) and per-sample loadings, `Lsyn96` is the number of synonymous sites per
trinucleotide context (dndscv RefCDS; `data/Lsyn.rds`, ≈ 23.3 M sites), and `g`
calibrates the mean per-sample synonymous burden to the **observed ≈ 64 synonymous
mutations per sample** in TCGA-LUAD (530 samples). Since the true `bmr`
(`sc = 359000`) already yields ≈ 63 syn/sample, **`g = 64.2 / 63 ≈ 1.02`** — the
silent matrix and the ERBB3 mutations share **one realistic LUAD-scale
background**. Both methods run on the **same simulated cohorts** as the
true-background result (the cohort RNG stream is frozen around the estimation step,
so per-iteration mutation counts are identical).

### Rerunning the estimated-background version

The estimated-background helpers are in
[`code/simulate_silent_estimate.R`](https://github.com/szhaolab/diffdriver-paper/blob/main/code/simulate_silent_estimate.R):
`build_silent_matrix()` (the LUAD-calibrated silent count matrix),
`coselens_estimated_rate()` (`N_syn/Lsyn96` per group → `dndscv_truerate`), and
`diffdriver_estimated_mr()` (fastTopics spectrum → per-site log-background). Per
iteration:

```{r eval=FALSE}
source("code/simulate_functions.R"); source("code/simulate_silent_estimate.R")
source("code/run_coselens.R"); source("code/dndscv_truerate.R")
library(diffdriver); library(coselens)
load("data/bmr_sim_inputs.rda"); load("data/hmm.rda")
load("data/codeSignature.rda");  load("data/hotspot1sig.rda")
cm <- readRDS("data/context96_map.rds"); Lsyn96 <- readRDS("data/Lsyn.rds")$Lsyn96

sim <- simulate_confounding(binary = TRUE, sganno = bmr_sim$sganno,
  sgmatrix = bmr_sim$sgmatrix, Nsample = 1500, para = c(0.5, 0.5),
  beta_gc = bmr_sim$beta_gc, signatures = bmr_sim$signature, loadings = bmr_sim$ll,
  rho = sqrt(0.78), hot = 1, hmm = hmm, sc = 359000)
bmr <- sim$bmrfold$bmr; e <- sim$pheno; g1 <- which(e == 1); g2 <- which(e == 0)
maf <- sim_to_maf(sim$mutlist, sim$annodata, e)

Y <- build_silent_matrix(bmr, Lsyn96, target_silent = 64.2)   # ~64 syn/sample

# coselens: estimated per-group rate from the silent counts
mr1 <- coselens_estimated_rate(Y, g1, Lsyn96, cm)
mr2 <- coselens_estimated_rate(Y, g2, Lsyn96, cm)
cos <- run_coselens(maf$group1, maf$group2, subset.genes.by = "ERBB3",
         true_rate = list(mutrates1 = mr1, mutrates2 = mr2, theta = 1e6),
         gene_list = "ERBB3", cv = NULL)$substitutions$pval

# DiffDriver: estimated per-site background from fastTopics(Y)
mr_est <- do.call(rbind, diffdriver_estimated_mr(Y, sim$bmrmtxlist, bmr))
dd <- ddmodel_binary(do.call(rbind, sim$mutlist), e, mr_est, sim$efsize$diffFe)$pvalue
```

The full benchmark loops this over `Nsample = 200 … 1500` (100 iterations each)
and stores the results in `data/pppower_estimatedBMR_bmr.Rd`.

### Estimated vs. true background

Both DiffDriver and Coselens **control the false positive rate whether they are
given the true background or estimate it** from the LUAD-calibrated silent count
matrix (all four curves stay near zero). The count-based benchmarks do **not** use
or estimate a background rate — they operate on the raw mutation counts, so the
"true vs estimated" distinction does not apply to them; they are shown only for
context (from the true-background result above) and rise toward ~0.6.

```{r estimated, fig.width=8, fig.height=6}
Nsim <- 100
SIZES <- c(200,400,600,800,1000,1200,1500)
load("data/pppower_estimatedBMR_bmr.Rd")     # object `estBMR`, keyed by N
load("data/pppower_mannwhitney_bmr.Rd")      # object `mw`

fp <- function(p) mean(p < 0.01, na.rm = TRUE)
rate <- function(field) sapply(as.character(SIZES), function(n) fp(estBMR[[n]][[field]]))

# background-modeling methods (this simulation): 4 series
cosT <- rate("coselens_true"); cosE <- rate("coselens_est")
ddT  <- rate("diff_true");     ddE  <- rate("diff_est")

# count-based benchmarks (for context; from the true-background sim, same setup)
lin <- fis <- bin <- lr <- mw_fp <- numeric(length(SIZES))
for (i in seq_along(SIZES)) {
  N <- SIZES[i]
  o <- new.env(); load(sprintf("data/bmr_archive/other/pppower_betaf0=1_sample%d_0.5_0.Rd", N), envir = o)
  s <- o$simuresother
  lin[i] <- fp(s$m1.pvalue); fis[i] <- fp(s$m2.pvalue); bin[i] <- fp(s$m3.pvalue); lr[i] <- fp(s$m4.pvalue)
  mw_fp[i] <- fp(mw[[as.character(N)]]$mannwhitney.pvalue)
}

x <- seq_along(SIZES)
par(mar = c(5,5,3,9), xpd = NA)
plot(x, ylim = c(0,0.7), col = "white", ylab = "False positive rate (p < 0.01)",
     xlab = "Number of samples", xaxt = "n", las = 1)
axis(1, at = x, labels = SIZES); grid(); abline(h = 0.01, lty = 3, col = "grey60")

# count-based (thin) for context
cb <- list(Linear = lin, Fisher = fis, Binomial = bin, LogisticR = lr, MannWhitney = mw_fp)
cbcol <- c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#1a9850")
for (j in seq_along(cb)) { lines(x, cb[[j]], col = cbcol[j], lwd = 1.2); points(x, cb[[j]], col = cbcol[j], pch = 20, cex = 0.9) }

# background-modeling: Coselens (orange) / DiffDriver (red); true = filled/solid, est = open/dashed
lines(x, cosT, col = "#ff7f00", lwd = 2);              points(x, cosT, col = "#ff7f00", pch = 19, cex = 1.4)
lines(x, cosE, col = "#ff7f00", lwd = 2, lty = 2);     points(x, cosE, col = "#ff7f00", pch = 1,  cex = 1.4, lwd = 2)
lines(x, ddT,  col = "#FF3D2E", lwd = 2);              points(x, ddT,  col = "#FF3D2E", pch = 17, cex = 1.4)
lines(x, ddE,  col = "#FF3D2E", lwd = 2, lty = 2);     points(x, ddE,  col = "#FF3D2E", pch = 2,  cex = 1.4, lwd = 2)

legend(par("usr")[2], par("usr")[4], bty = "n", cex = 0.85, xjust = 0,
  legend = c("Coselens (true bkg)","Coselens (estimated)",
             "DiffDriver (true bkg)","DiffDriver (estimated)","",
             "Linear","Fisher","Binomial","LogisticR","MannWhitney"),
  col = c("#ff7f00","#ff7f00","#FF3D2E","#FF3D2E",NA, cbcol),
  pch = c(19,1,17,2,NA, rep(20,5)),
  lty = c(1,2,1,2,NA, rep(1.2,5)), lwd = c(2,2,2,2,NA, rep(1.2,5)))
title("Estimated vs. true background under signature confounding")
```

The two dashed curves (estimated background) sit right on top of their solid
counterparts (true background), near zero — estimating the background from a
realistic, LUAD-calibrated silent count matrix does not reintroduce the
confounding, for either DiffDriver or Coselens.
