---
title: "Simulation: Confounding Between Phenotype and Mutational Signatures"
author: "Jie Zhou, Siming Zhao"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: hide
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This simulation demonstrates that when the phenotype is correlated with
mutational signatures — but **not** with selection — benchmark methods produce
false positives while DiffDriver correctly controls the false positive rate.

### Simulation setup

- **Background mutation rate**: Estimated from LUAD mutation data (527 samples,
  bootstrapped up to 1,500 samples). The per-sample background rate is driven by
  the real per-sample signature loadings (`ll$LUAD`), which are a composition
  (constant column sums), so a summed-loading covariate cannot adjust the
  confounding away.
- **Mutational signatures**: Six signatures extracted from LUAD using
  [fastTopics](https://stephenslab.github.io/fastTopics/). Signature 3 — a
  T>G–dominated (smoking-like) signature whose highest-loading context is
  `C[T>G]G` (nucleotide type 34) — is selected as the confounding signature.
- **Phenotype**: Simulated to correlate with signature 3 ($R^2 = 0.78$), but
  with **zero correlation** with selection (`para = c(0.5, 0.5)`).
- **Mutations**: Simulated for a single gene (ERBB3) using oncogene mutation
  rates (background + selection), with two hotspots placed at nucleotide type 34
  (the context with the highest loading in signature 3), so that the
  hotspot-dominated mutations track the phenotype.

### Key result

Because the phenotype correlates with a mutational signature that affects
overall mutation rates, benchmark methods (Linear regression, Fisher's exact
test, Binomial test, Logistic regression, and the Mann-Whitney rank-sum test)
incorrectly detect differential selection. DiffDriver, which accounts for
signature-based background mutation rates, maintains a low false positive rate.

## Rerunning the simulation

The simulation functions live in [`code/simulate_functions.R`](https://github.com/szhaolab/diffdriver-paper/blob/main/code/simulate_functions.R)
under the "CONFOUNDING scenario" section (`build_confounded_bmr`,
`simulate_confounding`, `run_confounding_benchmarks`,
`run_confounding_diffdriver`, `run_confounding_diffdriver_fixed`). The benchmark
methods are all in `run_confounding_benchmarks`
(Linear/Fisher/Binomial/LogisticR + Mann-Whitney); DiffDriver is run by
`run_confounding_diffdriver_fixed`.

The inputs are bundled in `data/bmr_sim_inputs.rda` (`beta_gc`, `beta_gcFix`,
`signature`, `ll`, `sganno`, `sgmatrix`), together with `data/hmm.rda`,
`data/codeSignature.rda`, and `data/hotspot1sig.rda` (hotspots at nttype 34).
A single sample size is reproduced with:

```{r eval=FALSE}
library(diffdriver); library(data.table); library(Matrix)
source("code/simulate_functions.R")
load("data/bmr_sim_inputs.rda"); load("data/hmm.rda")
load("data/codeSignature.rda");  load("data/hotspot1sig.rda")

set.seed(1)
res_other <- run_confounding_benchmarks(
  binary = TRUE, Niter = 100,
  sganno = bmr_sim$sganno, sgmatrix = bmr_sim$sgmatrix,
  Nsample = 1500, para = c(0.5, 0.5),      # no differential selection
  rho = sqrt(0.78),                        # phenotype ~ signature 3, R^2 = 0.78
  signatures = bmr_sim$signature, loadings = bmr_sim$ll,
  beta_gc = bmr_sim$beta_gc, hot = 1, hmm = hmm,
  sc = 359000)                             # background-rate scale (~188 mutations)

# false-positive rate per method (p < 0.01), no true signal so all are FPs
sapply(res_other[c("m1.pvalue","m2.pvalue","m3.pvalue","m4.pvalue",
                    "mannwhitney.pvalue")], function(p) mean(p < 0.01))
```

The full benchmark loops this over `Nsample = 200 … 1500`; the Mann-Whitney
p-values used below were produced this way and saved to
`data/pppower_mannwhitney_bmr.Rd`. Because the original
`scripts/simulate_functions.R` (and the exact HMM coefficients) were lost, the
rerun is *close* to — but not bit-identical to — the archived runs; the Linear
and Fisher/Binomial/LogisticR curves plotted below are read from the original
archived results.



```{r}
Nsim=100
SIZES <- c(200,400,600,800,1000,1200,1500)

# Mann-Whitney FPs from the rerun (code/simulate_functions.R::run_confounding_benchmarks)
load("data/pppower_mannwhitney_bmr.Rd")   # object `mw`, keyed by sample size

mylist <- list()
for (i1 in c(1)){
   for (rho in c(0)) {
    for (i4 in c(0.5)) {
    n<-0
    for(i3 in SIZES){
      n <- n+1
      load(paste0("data/bmr_archive/diff/pppower_betaf0=",i1, "_sample",i3,"_",i4,"_",rho,".Rd"))
         load(paste0("data/bmr_archive/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i4,"_", rho,".Rd"))
          load(paste0("data/bmr_archive/other/pppower_betaf0=",i1,"_sample",i3,"_",i4,"_",rho,".Rd"))

      m5.pvalue <- simuresdiff[["m1.pvalue"]]
       m6.pvalue <- simuresdiffFix[["m1.pvalue"]]
      m1.pvalue <- simuresother[["m1.pvalue"]]
      m2.pvalue <- simuresother[["m2.pvalue"]]
      m3.pvalue <- simuresother[["m3.pvalue"]]
      m4.pvalue <- simuresother[["m4.pvalue"]]
      m7.pvalue <- mw[[as.character(i3)]][["mannwhitney.pvalue"]]   # Mann-Whitney
     mylist[[n]] <- c(length(m1.pvalue[m1.pvalue <0.01]),length(m2.pvalue[m2.pvalue <0.01]),length(m3.pvalue[m3.pvalue <0.01]), length(m4.pvalue[m4.pvalue <0.01]),length(m7.pvalue[m7.pvalue <0.01]),length(m5.pvalue[m5.pvalue <0.01]),length(m6.pvalue[m6.pvalue <0.01]))/Nsim
    }
    outdt <- do.call(cbind, mylist)
    rownames(outdt)=c("Linear","Fisher","Binomial","LogisticR","MannWhitney","diffDriver","avdiffDriver")
    colnames(outdt) <- SIZES
    # colors: Linear, Fisher, Binomial, LogisticR, MannWhitney, diff, diffFix
    mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#1a9850","#FF3D2E","#f781bf")
    mypch=c(19,19,19,19,3,17,17)
     plot(1:dim(outdt)[2],ylim=c(0,1), col="white", ylab = "False positive rate (p < 0.01)", xlab="Number of samples", yaxt="n", xaxt="n")
     title( paste( "SC(scale efficient)=",i1,",","para=(",i4,1-i4,")", ",", sep=" "))
    yticks_val <- seq(0,1,0.1)
    axis(2, at=yticks_val, lab=paste0(yticks_val),las=2)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt))
    grid()
    for (r in 1:7){
      points(outdt[r,],col=mycol[r], pch=mypch[r], cex=1.3)
      lines(outdt[r,],col=mycol[r])
    }
    legend("topleft", legend = c("Linear","Fisher","Binomial","LogisticR","MannWhitney","diff","diffFix"), col= mycol, lwd=1, bty="n", lty=rep(NA,7), pch=mypch)
  }
}
  }

```
