---
title: "Simulation: Confounding Between Phenotype and Mutational Signatures"
author: "Jie Zhou, Siming Zhao"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: hide
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This simulation demonstrates that when the phenotype is correlated with
mutational signatures — but **not** with selection — benchmark methods produce
false positives while DiffDriver correctly controls the false positive rate.

### Simulation setup

- **Background mutation rate**: Estimated from LUAD mutation data (527 samples,
  bootstrapped up to 1,500 samples).
- **Mutational signatures**: Six signatures extracted from LUAD using
  [fastTopics](https://stephenslab.github.io/fastTopics/). Signature 3 is
  selected as the confounding signature.
- **Phenotype**: Simulated to correlate with signature 3 ($R^2 = 0.78$), but
  with **zero correlation** with selection.
- **Mutations**: Simulated for a single gene using oncogene mutation rates
  (background + selection), with two hotspots placed at nucleotide type 63
  (the type with the highest loading in signature 3).

### Key result

Because the phenotype correlates with a mutational signature that affects
overall mutation rates, benchmark methods (Linear regression, Fisher's exact
test, Binomial test, Logistic regression) incorrectly detect differential
selection. DiffDriver, which accounts for signature-based background mutation
rates, maintains a low false positive rate.  



```{r}
Nsim=100
mylist <- list()
for (i1 in c(1)){
   for (rho in c(0)) {
    for (i4 in c(0.5)) {
    n<-0
    for(i3 in c(200,400,600,800,1000,1200,1500)){
      n <- n+1
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/randombmrsingle/tsgsimple_LUAD_fixed_63/diff/pppower_betaf0=",i1, "_sample",i3,"_",i4,"_",rho,".Rd"))
         load(paste0("/Volumes/Szhao/jiezhou/diffdata/randombmrsingle/tsgsimple_LUAD_fixed_63/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i4,"_", rho,".Rd"))
          load(paste0("/Volumes/Szhao/jiezhou/diffdata/randombmrsingle/tsgsimple_LUAD_fixed_63/other/pppower_betaf0=",i1,"_sample",i3,"_",i4,"_",rho,".Rd"))
 
      m5.pvalue <- simuresdiff[["m1.pvalue"]]
       m6.pvalue <- simuresdiffFix[["m1.pvalue"]]
      m1.pvalue <- simuresother[["m1.pvalue"]]
      m2.pvalue <- simuresother[["m2.pvalue"]]
      m3.pvalue <- simuresother[["m3.pvalue"]]
      m4.pvalue <- simuresother[["m4.pvalue"]]
    #  mutNumber1[m,n]=mean(simuresbase$`#mut`)/i3*100
     mylist[[n]] <- c(length(m1.pvalue[m1.pvalue <0.01]),length(m2.pvalue[m2.pvalue <0.01]),length(m3.pvalue[m3.pvalue <0.01]), length(m4.pvalue[m4.pvalue <0.01]),length(m5.pvalue[m5.pvalue <0.01]),length(m6.pvalue[m6.pvalue <0.01]))/Nsim
    }
    outdt <- do.call(cbind, mylist)
    rownames(outdt)=c("Linear","Fisher","Binomial","LogisticR","diffDriver","avdiffDriver")
    colnames(outdt) <- c(200,400,600,800,1000,1200,1500)
    mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#fdae61","#FF3D2E","#FFFF00", "#2b2930")
 # mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#fdae61","#FF3D2E")
     plot(1:dim(outdt)[2],ylim=c(0,1), col="white", ylab = "Power", xlab="Number of samples", yaxt="n", xaxt="n")
      # title(paste("f0=",i1,",", "beta_gc=(",i2,",","1.513", ")", "P(E=1) =0.5", "P(positive selection|E=1)=0.8", "P(positive selection|E=1)=0.2", sep=" "))
     title( paste( "SC(scale efficient)=",i1,",","para=(",i4,1-i4,")", ",", sep=" "))
    #yticks_val <- pretty(outdt)
    yticks_val <- seq(0,1,0.1)
    #axis(2, lab=paste0(yticks_val/Nsim * 100, " %"),las=2)
    axis(2, at=yticks_val, lab=paste0(yticks_val),las=2)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt))
    grid()
    for (r in 1:6){
      points(outdt[r,],col=mycol[r+1], pch=19, cex=1.3)
      lines(outdt[r,],col=mycol[r+1])
    }
    legend("topleft", legend = c("Linear","Fisher","Binomial","LogisticR","diff","diffFix"), col= mycol[2:7], lwd=1, bty="n", lty=rep(NA,6), pch=rep(19,6))
  }
}
  }

```
