---
title: "Simulation Power Results"
author: "Jie Zhou, Siming Zhao"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

For 200 genes with five functional annotations, we compare the power (sensitivity)
and false positive rate (FPR) of DiffDriver against benchmark methods: Linear
regression, Fisher's exact test, Binomial test, and Logistic Regression.

The first 50 genes have differential selection while the remaining 150 do not.
FDR is controlled at 0.1 using the BH procedure.

### Simulation parameters

These simulations use the same pipeline described in
[Simulation Code](simulation_code.html). The key parameters are:

| Parameter | Description |
|-----------|-------------|
| `betaf0`  | Baseline functional effect size: 0 or 1 |
| `para`    | Binary: selection probabilities c(prob_case, prob_control). Continuous: c(mean, sd, intercept, slope) |
| `beta_gc` | Functional effect coefficients (TSGpars or OGpars) |
| `Nsample` | Varies: 200, 400, 600, 800, 1000, 1200, 1500 |
| `Niter`   | 100 (binary) or 50 (continuous) per gene |

### Functional effect coefficients

- **TSG parameters**: `(Intercept)=0.851, functypecode8=0.458, mycons=0.137, sift=0.124, phylop100=0.083, MA=0.306`
- **OG parameters**: `(Intercept)=0.450, functypecode8=-0.186, mycons=0.316, sift=0.020, phylop100=0.256, MA=0.140`

---

## TSG, Binary Phenotype

### para = c(0.8, 0.2)

```{r tsg-binary-08}
Nsim <- 100
rr <- 0.1
mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61","#FFFF00", "#2b2930")
methods <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
sample_sizes <- c(200, 400, 600, 800, 1000, 1200, 1500)
mylist1 <- mylist2 <- replicate(length(sample_sizes), replicate(length(methods), c(), simplify = FALSE), simplify = FALSE)
m.pvalue <- lapply(1:length(methods), function(x) matrix(nrow = 200, ncol = Nsim))

for (i1 in 0) {
  n <- 0
  for (i3 in sample_sizes) {
    n <- n + 1
    for (i2 in 1:200) {
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/diff/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/other/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/diffFixFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))

      m.pvalue[[1]][i2, ] <- simuresdiffFix[["m1.pvalue"]]
      m.pvalue[[2]][i2, ] <- simuresother[["m1.pvalue"]]
      m.pvalue[[3]][i2, ] <- simuresother[["m2.pvalue"]]
      m.pvalue[[4]][i2, ] <- simuresother[["m3.pvalue"]]
      m.pvalue[[5]][i2, ] <- simuresother[["m4.pvalue"]]
    }
    for (j2 in 1:Nsim) {
      for (k in 1:length(methods)) {
        index1 <- which(p.adjust(m.pvalue[[k]][, j2], method = "BH") < rr)
        sensitivity <- ifelse(length(index1) != 0, length(which(index1 < 51)) / 50, 0)
        specificity <- ifelse(length(index1) != 0, length(which(index1 > 50)) / length(index1), 0)
        mylist1[[n]][[k]] <- c(sensitivity, mylist1[[n]][[k]])
        mylist2[[n]][[k]] <- c(specificity, mylist2[[n]][[k]])
      }
    }
  }
}

par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))
for (i in 1:2) {
  mylist <- if (i == 1) mylist1 else mylist2
  outdt <- sapply(mylist, function(x) sapply(x, mean))
  rownames(outdt) <- methods
  colnames(outdt) <- sample_sizes
  plot(1:dim(outdt)[2], ylim = c(0, ifelse(i == 1, 0.4, 1)), col = "white",
       ylab = ifelse(i == 1, "Power", "FPR"),
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = 'n',
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title(paste("para =", ifelse(i == 1, "(0.8, 0.2)", "(0.5, 0.5), FDR = 0.1")))
  yticks_val <- seq(0, ifelse(i == 1, 0.4, 1), 0.1)
  axis(2, at = yticks_val, lab = paste0(yticks_val * 100, "%"), las = 2, cex.axis = 1.3)
  axis(1, at = 1:dim(outdt)[2], lab = colnames(outdt), cex.axis = 1.3)
  grid()
  for (r in 1:nrow(outdt)) {
    points(1:dim(outdt)[2], outdt[r, ], col = mycol[r], pch = 19, cex = 1.8)
    lines(1:dim(outdt)[2], outdt[r, ], col = mycol[r], lwd = 1.5)
  }
  if (i == 1) {
    legend("topleft", legend = methods, col = mycol[1:length(methods)],
           lwd = 1, bty = "n", lty = rep(NA, 5), pch = rep(19, 5))
  }
}
```

### para = c(0.9, 0.1)

```{r tsg-binary-09}
Nsim <- 100
rr <- 0.1
mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61","#FFFF00", "#2b2930")
methods <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
sample_sizes <- c(200, 400, 600, 800, 1000, 1200, 1500)
mylist1 <- mylist2 <- replicate(length(sample_sizes), replicate(length(methods), c(), simplify = FALSE), simplify = FALSE)
m.pvalue <- lapply(1:length(methods), function(x) matrix(nrow = 200, ncol = Nsim))

for (i1 in 0) {
  n <- 0
  for (i3 in sample_sizes) {
    n <- n + 1
    for (i2 in 1:200) {
     load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/TSG0.9/TSG_gene/diff/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
     load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/TSG0.9/TSG_gene/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
     load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/TSG0.9/TSG_gene/other/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
     load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/TSG0.9/TSG_gene/diffFixFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))

      m.pvalue[[1]][i2, ] <- simuresdiffFix[["m1.pvalue"]]
      m.pvalue[[2]][i2, ] <- simuresother[["m1.pvalue"]]
      m.pvalue[[3]][i2, ] <- simuresother[["m2.pvalue"]]
      m.pvalue[[4]][i2, ] <- simuresother[["m3.pvalue"]]
      m.pvalue[[5]][i2, ] <- simuresother[["m4.pvalue"]]
    }
    for (j2 in 1:Nsim) {
      for (k in 1:length(methods)) {
        index1 <- which(p.adjust(m.pvalue[[k]][, j2], method = "BH") < rr)
        sensitivity <- ifelse(length(index1) != 0, length(which(index1 < 51)) / 50, 0)
        specificity <- ifelse(length(index1) != 0, length(which(index1 > 50)) / length(index1), 0)
        mylist1[[n]][[k]] <- c(sensitivity, mylist1[[n]][[k]])
        mylist2[[n]][[k]] <- c(specificity, mylist2[[n]][[k]])
      }
    }
  }
}

par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))
for (i in 1:2) {
  mylist <- if (i == 1) mylist1 else mylist2
  outdt <- sapply(mylist, function(x) sapply(x, mean))
  rownames(outdt) <- methods
  colnames(outdt) <- sample_sizes
  plot(1:dim(outdt)[2], ylim = c(0, ifelse(i == 1, 0.4, 1)), col = "white",
       ylab = ifelse(i == 1, "Power", "FPR"),
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = 'n',
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title(paste("para =", ifelse(i == 1, "(0.9, 0.1)", "(0.5, 0.5), FDR = 0.1")))
  yticks_val <- seq(0, ifelse(i == 1, 0.4, 1), 0.1)
  axis(2, at = yticks_val, lab = paste0(yticks_val * 100, "%"), las = 2, cex.axis = 1.3)
  axis(1, at = 1:dim(outdt)[2], lab = colnames(outdt), cex.axis = 1.3)
  grid()
  for (r in 1:nrow(outdt)) {
    points(1:dim(outdt)[2], outdt[r, ], col = mycol[r], pch = 19, cex = 1.8)
    lines(1:dim(outdt)[2], outdt[r, ], col = mycol[r], lwd = 1.5)
  }
}
```

---

## TSG, Continuous Phenotype

Continuous phenotype with `para = c(0, 2, 0, 1)`: phenotype ~ N(0, 2), selection
probability determined by logistic regression with intercept = 0 and slope = 1.

### betaf0 = 0

```{r tsg-conti-f0}
Nsim <- 50
rr <- 0.1
methods <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
sample_sizes <- c(200, 400, 600, 800, 1000, 1200)
mycol <- c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61")
mylist1 <- mylist2 <- list()
m.pvalue <- lapply(1:6, function(x) matrix(nrow = 200, ncol = Nsim))

for (i1 in c(0)) {
  n <- 0
  for (i3 in sample_sizes) {
    n <- n + 1
    sensitivity1=c(); specificity1=c()
    sensitivity2=c(); specificity2=c()
    sensitivity3=c(); specificity3=c()
    sensitivity4=c(); specificity4=c()
    sensitivity5=c(); specificity5=c()
    sensitivity6=c(); specificity6=c()

    for (i2 in 1:200) {
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2/diff/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2/other/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))

      m.pvalue[[1]][i2,] <- simuresother[["m1.pvalue"]]
      m.pvalue[[2]][i2,] <- simuresother[["m2.pvalue"]]
      m.pvalue[[3]][i2,] <- simuresother[["m3.pvalue"]]
      m.pvalue[[4]][i2,] <- simuresother[["m4.pvalue"]]
      m.pvalue[[5]][i2,] <- simuresdiffFix[["m1.pvalue"]]
      m.pvalue[[6]][i2,] <- simuresdiff[["m1.pvalue"]]
    }
    for (j2 in 1:Nsim) {
      for (k in 1:5) {
        idx <- k
        if (k == 1) idx <- 5  # DiffDriver
        if (k == 2) idx <- 1  # Linear
        if (k == 3) idx <- 2  # Fisher
        if (k == 4) idx <- 3  # Binomial
        if (k == 5) idx <- 4  # LogisticR
        index1 <- which(p.adjust(m.pvalue[[idx]][, j2], method = "BH") < rr)
        sens <- ifelse(length(index1) != 0, length(which(index1 < 51)) / 50, 0)
        spec <- ifelse(length(index1) != 0, length(which(index1 > 50)) / length(index1), 0)
        if (k == 1) { sensitivity5 <- c(sensitivity5, sens); specificity5 <- c(specificity5, spec) }
        if (k == 2) { sensitivity1 <- c(sensitivity1, sens); specificity1 <- c(specificity1, spec) }
        if (k == 3) { sensitivity2 <- c(sensitivity2, sens); specificity2 <- c(specificity2, spec) }
        if (k == 4) { sensitivity3 <- c(sensitivity3, sens); specificity3 <- c(specificity3, spec) }
        if (k == 5) { sensitivity4 <- c(sensitivity4, sens); specificity4 <- c(specificity4, spec) }
      }
    }
    mylist1[[n]] <- c(mean(sensitivity5), mean(sensitivity1), mean(sensitivity2), mean(sensitivity3), mean(sensitivity4))
    mylist2[[n]] <- c(mean(specificity5), mean(specificity1), mean(specificity2), mean(specificity3), mean(specificity4))
  }

  outdt <- do.call(cbind, mylist1)
  rownames(outdt) <- methods
  colnames(outdt) <- sample_sizes
  x <- 1:dim(outdt)[2]

  par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))
  plot(x, ylim = c(0, 0.6), col = "white", ylab = "Power",
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = "n",
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title(paste("betaf0 =", i1, ", para = (0, 2, 0, 1)"))
  axis(2, at = seq(0, 0.6, 0.1), lab = paste0(seq(0, 60, 10), "%"), las = 2, cex.axis = 1.3)
  axis(1, at = x, lab = colnames(outdt), cex.axis = 1.3)
  grid()
  for (r in 1:5) {
    points(x, outdt[r, ], col = mycol[r], pch = 19, cex = 1.8)
    lines(x, outdt[r, ], col = mycol[r], lwd = 1.5)
  }
  legend("topleft", legend = methods, col = mycol[1:5], lwd = 1, bty = "n",
         lty = rep(NA, 5), pch = rep(19, 5))

  outdt2 <- do.call(cbind, mylist2)
  rownames(outdt2) <- methods
  colnames(outdt2) <- sample_sizes
  plot(x, ylim = c(0, 1), col = "white", ylab = "FPR",
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = "n",
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title("para = (0, 2, 0, 0), FDR = 0.1")
  axis(2, at = seq(0, 1, 0.1), lab = paste0(seq(0, 100, 10), "%"), las = 2, cex.axis = 1.3)
  axis(1, at = x, lab = colnames(outdt2), cex.axis = 1.3)
  grid()
  for (r in 1:5) {
    points(x, outdt2[r, ], col = mycol[r], pch = 19, cex = 1.8)
    lines(x, outdt2[r, ], col = mycol[r], lwd = 1.5)
  }
}
```

### betaf0 = 1

```{r tsg-conti-f1}
Nsim <- 50
rr <- 0.1
methods <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
sample_sizes <- c(200, 400, 600, 800, 1000, 1200)
mycol <- c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61")
mylist1 <- mylist2 <- list()
m.pvalue <- lapply(1:6, function(x) matrix(nrow = 200, ncol = Nsim))

for (i1 in c(1)) {
  n <- 0
  for (i3 in sample_sizes) {
    n <- n + 1
    sensitivity1=c(); specificity1=c()
    sensitivity2=c(); specificity2=c()
    sensitivity3=c(); specificity3=c()
    sensitivity4=c(); specificity4=c()
    sensitivity5=c(); specificity5=c()
    sensitivity6=c(); specificity6=c()

    for (i2 in 1:200) {
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2/diff/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2/other/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))

      m.pvalue[[1]][i2,] <- simuresother[["m1.pvalue"]]
      m.pvalue[[2]][i2,] <- simuresother[["m2.pvalue"]]
      m.pvalue[[3]][i2,] <- simuresother[["m3.pvalue"]]
      m.pvalue[[4]][i2,] <- simuresother[["m4.pvalue"]]
      m.pvalue[[5]][i2,] <- simuresdiffFix[["m1.pvalue"]]
      m.pvalue[[6]][i2,] <- simuresdiff[["m1.pvalue"]]
    }
    for (j2 in 1:Nsim) {
      for (k in 1:5) {
        idx <- k
        if (k == 1) idx <- 5
        if (k == 2) idx <- 1
        if (k == 3) idx <- 2
        if (k == 4) idx <- 3
        if (k == 5) idx <- 4
        index1 <- which(p.adjust(m.pvalue[[idx]][, j2], method = "BH") < rr)
        sens <- ifelse(length(index1) != 0, length(which(index1 < 51)) / 50, 0)
        spec <- ifelse(length(index1) != 0, length(which(index1 > 50)) / length(index1), 0)
        if (k == 1) { sensitivity5 <- c(sensitivity5, sens); specificity5 <- c(specificity5, spec) }
        if (k == 2) { sensitivity1 <- c(sensitivity1, sens); specificity1 <- c(specificity1, spec) }
        if (k == 3) { sensitivity2 <- c(sensitivity2, sens); specificity2 <- c(specificity2, spec) }
        if (k == 4) { sensitivity3 <- c(sensitivity3, sens); specificity3 <- c(specificity3, spec) }
        if (k == 5) { sensitivity4 <- c(sensitivity4, sens); specificity4 <- c(specificity4, spec) }
      }
    }
    mylist1[[n]] <- c(mean(sensitivity5), mean(sensitivity1), mean(sensitivity2), mean(sensitivity3), mean(sensitivity4))
    mylist2[[n]] <- c(mean(specificity5), mean(specificity1), mean(specificity2), mean(specificity3), mean(specificity4))
  }

  outdt <- do.call(cbind, mylist1)
  rownames(outdt) <- methods
  colnames(outdt) <- sample_sizes
  x <- 1:dim(outdt)[2]

  par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))
  plot(x, ylim = c(0, 0.6), col = "white", ylab = "Power",
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = "n",
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title(paste("betaf0 =", i1, ", para = (0, 2, 0, 2)"))
  axis(2, at = seq(0, 1, 0.1), lab = paste0(seq(0, 100, 10), "%"), las = 2, cex.axis = 1.3)
  axis(1, at = x, lab = colnames(outdt), cex.axis = 1.3)
  grid()
  for (r in 1:5) {
    points(x, outdt[r, ], col = mycol[r], pch = 19, cex = 1.8)
    lines(x, outdt[r, ], col = mycol[r], lwd = 1.5)
  }

  outdt2 <- do.call(cbind, mylist2)
  rownames(outdt2) <- methods
  colnames(outdt2) <- sample_sizes
  plot(x, ylim = c(0, 1), col = "white", ylab = "FPR",
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = "n",
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title("para = (0, 2, 0, 0), FDR = 0.1")
  axis(2, at = seq(0, 1, 0.1), lab = paste0(seq(0, 100, 10), "%"), las = 2, cex.axis = 1.3)
  axis(1, at = x, lab = colnames(outdt2), cex.axis = 1.3)
  grid()
  for (r in 1:5) {
    points(x, outdt2[r, ], col = mycol[r], pch = 19, cex = 1.8)
    lines(x, outdt2[r, ], col = mycol[r], lwd = 1.5)
  }
}
```

---

## OG, Binary Phenotype

### para = c(0.8, 0.2)

```{r og-binary-08}
Nsim <- 100
rr <- 0.1
mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61","#FFFF00", "#2b2930")
methods <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
sample_sizes <- c(200, 400, 600, 800, 1000, 1200, 1500)
mylist1 <- mylist2 <- replicate(length(sample_sizes), replicate(length(methods), c(), simplify = FALSE), simplify = FALSE)
m.pvalue <- lapply(1:length(methods), function(x) matrix(nrow = 200, ncol = Nsim))

for (i1 in 0.8) {
  n <- 0
  for (i3 in sample_sizes) {
    n <- n + 1
    for (i2 in 1:200) {
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/binaryEM/OG_gene/diff/pppower_betaf0=0_sample",i3,"_",i1,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/binaryEM/OG_gene/diffFix/pppower_betaf0=0_sample",i3,"_",i1,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/binaryEM/OG_gene/other/pppower_betaf0=0_sample",i3,"_",i1,"_",i2,".Rd"))

      m.pvalue[[1]][i2, ] <- simuresdiffFix[["m1.pvalue"]]
      m.pvalue[[2]][i2, ] <- simuresother[["m1.pvalue"]]
      m.pvalue[[3]][i2, ] <- simuresother[["m2.pvalue"]]
      m.pvalue[[4]][i2, ] <- simuresother[["m3.pvalue"]]
      m.pvalue[[5]][i2, ] <- simuresother[["m4.pvalue"]]
    }
    for (j2 in 1:Nsim) {
      for (k in 1:length(methods)) {
        index1 <- which(p.adjust(m.pvalue[[k]][, j2], method = "BH") < rr)
        sensitivity <- ifelse(length(index1) != 0, length(which(index1 < 51)) / 50, 0)
        specificity <- ifelse(length(index1) != 0, length(which(index1 > 50)) / length(index1), 0)
        mylist1[[n]][[k]] <- c(sensitivity, mylist1[[n]][[k]])
        mylist2[[n]][[k]] <- c(specificity, mylist2[[n]][[k]])
      }
    }
  }
}

par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))
for (i in 1:2) {
  mylist <- if (i == 1) mylist1 else mylist2
  outdt <- sapply(mylist, function(x) sapply(x, mean))
  rownames(outdt) <- methods
  colnames(outdt) <- sample_sizes
  plot(1:dim(outdt)[2], ylim = c(0, ifelse(i == 1, 0.5, 1)), col = "white",
       ylab = ifelse(i == 1, "Power", "FPR"),
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = 'n',
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title(paste("para =", ifelse(i == 1, "(0.8, 0.2)", "(0.5, 0.5), FDR = 0.1")))
  yticks_val <- seq(0, ifelse(i == 1, 0.5, 1), 0.1)
  axis(2, at = yticks_val, lab = paste0(yticks_val * 100, "%"), las = 2, cex.axis = 1.3)
  axis(1, at = 1:dim(outdt)[2], lab = colnames(outdt), cex.axis = 1.3)
  grid()
  for (r in 1:nrow(outdt)) {
    points(1:dim(outdt)[2], outdt[r, ], col = mycol[r], pch = 19, cex = 1.8)
    lines(1:dim(outdt)[2], outdt[r, ], col = mycol[r], lwd = 1.5)
  }
  if (i == 1) {
    legend("topleft", legend = methods, col = mycol[1:length(methods)],
           lwd = 1, bty = "n", lty = NA, pch = 19)
  }
}
```

### para = c(0.9, 0.1)

```{r og-binary-09}
Nsim <- 100
rr <- 0.1
mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61","#FFFF00", "#2b2930")
methods <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
sample_sizes <- c(200, 400, 600, 800, 1000, 1200, 1500)
mylist1 <- mylist2 <- replicate(length(sample_sizes), replicate(length(methods), c(), simplify = FALSE), simplify = FALSE)
m.pvalue <- lapply(1:length(methods), function(x) matrix(nrow = 200, ncol = Nsim))

for (i1 in 0.9) {
  n <- 0
  for (i3 in sample_sizes) {
    n <- n + 1
    for (i2 in 1:200) {
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/binaryEM/OG_gene/diff/pppower_betaf0=0_sample",i3,"_",i1,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/binaryEM/OG_gene/diffFix/pppower_betaf0=0_sample",i3,"_",i1,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/binaryEM/OG_gene/other/pppower_betaf0=0_sample",i3,"_",i1,"_",i2,".Rd"))

      m.pvalue[[1]][i2, ] <- simuresdiffFix[["m1.pvalue"]]
      m.pvalue[[2]][i2, ] <- simuresother[["m1.pvalue"]]
      m.pvalue[[3]][i2, ] <- simuresother[["m2.pvalue"]]
      m.pvalue[[4]][i2, ] <- simuresother[["m3.pvalue"]]
      m.pvalue[[5]][i2, ] <- simuresother[["m4.pvalue"]]
    }
    for (j2 in 1:Nsim) {
      for (k in 1:length(methods)) {
        index1 <- which(p.adjust(m.pvalue[[k]][, j2], method = "BH") < rr)
        sensitivity <- ifelse(length(index1) != 0, length(which(index1 < 51)) / 50, 0)
        specificity <- ifelse(length(index1) != 0, length(which(index1 > 50)) / length(index1), 0)
        mylist1[[n]][[k]] <- c(sensitivity, mylist1[[n]][[k]])
        mylist2[[n]][[k]] <- c(specificity, mylist2[[n]][[k]])
      }
    }
  }
}

par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))
for (i in 1:2) {
  mylist <- if (i == 1) mylist1 else mylist2
  outdt <- sapply(mylist, function(x) sapply(x, mean))
  rownames(outdt) <- methods
  colnames(outdt) <- sample_sizes
  plot(1:dim(outdt)[2], ylim = c(0, ifelse(i == 1, 0.5, 1)), col = "white",
       ylab = ifelse(i == 1, "Power", "FPR"),
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = 'n',
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title(paste("para =", ifelse(i == 1, "(0.9, 0.1)", "(0.5, 0.5), FDR = 0.1")))
  yticks_val <- seq(0, ifelse(i == 1, 0.5, 1), 0.1)
  axis(2, at = yticks_val, lab = paste0(yticks_val * 100, "%"), las = 2, cex.axis = 1.3)
  axis(1, at = 1:dim(outdt)[2], lab = colnames(outdt), cex.axis = 1.3)
  grid()
  for (r in 1:nrow(outdt)) {
    points(1:dim(outdt)[2], outdt[r, ], col = mycol[r], pch = 19, cex = 1.8)
    lines(1:dim(outdt)[2], outdt[r, ], col = mycol[r], lwd = 1.5)
  }
}
```

---

## OG, Continuous Phenotype

Continuous phenotype with `para = c(0, 2, 0, 1)`: phenotype ~ N(0, 2), selection
probability determined by logistic regression with intercept = 0 and slope = 1.

### betaf0 = 0

```{r og-conti-f0}
Nsim <- 50
rr <- 0.1
methods <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
sample_sizes <- c(200, 400, 600, 800, 1000, 1200)
mycol <- c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61")
mylist1 <- mylist2 <- list()
m.pvalue <- lapply(1:6, function(x) matrix(nrow = 200, ncol = Nsim))

for (i1 in c(0)) {
  n <- 0
  for (i3 in sample_sizes) {
    n <- n + 1
    sensitivity1=c(); specificity1=c()
    sensitivity2=c(); specificity2=c()
    sensitivity3=c(); specificity3=c()
    sensitivity4=c(); specificity4=c()
    sensitivity5=c(); specificity5=c()
    sensitivity6=c(); specificity6=c()

    for (i2 in 1:200) {
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene/diff/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene/other/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))

      m.pvalue[[1]][i2,] <- simuresother[["m1.pvalue"]]
      m.pvalue[[2]][i2,] <- simuresother[["m2.pvalue"]]
      m.pvalue[[3]][i2,] <- simuresother[["m3.pvalue"]]
      m.pvalue[[4]][i2,] <- simuresother[["m4.pvalue"]]
      m.pvalue[[5]][i2,] <- simuresdiffFix[["m1.pvalue"]]
      m.pvalue[[6]][i2,] <- simuresdiff[["m1.pvalue"]]
    }
    for (j2 in 1:Nsim) {
      for (k in 1:5) {
        idx <- k
        if (k == 1) idx <- 5
        if (k == 2) idx <- 1
        if (k == 3) idx <- 2
        if (k == 4) idx <- 3
        if (k == 5) idx <- 4
        index1 <- which(p.adjust(m.pvalue[[idx]][, j2], method = "BH") < rr)
        sens <- ifelse(length(index1) != 0, length(which(index1 < 51)) / 50, 0)
        spec <- ifelse(length(index1) != 0, length(which(index1 > 50)) / length(index1), 0)
        if (k == 1) { sensitivity5 <- c(sensitivity5, sens); specificity5 <- c(specificity5, spec) }
        if (k == 2) { sensitivity1 <- c(sensitivity1, sens); specificity1 <- c(specificity1, spec) }
        if (k == 3) { sensitivity2 <- c(sensitivity2, sens); specificity2 <- c(specificity2, spec) }
        if (k == 4) { sensitivity3 <- c(sensitivity3, sens); specificity3 <- c(specificity3, spec) }
        if (k == 5) { sensitivity4 <- c(sensitivity4, sens); specificity4 <- c(specificity4, spec) }
      }
    }
    mylist1[[n]] <- c(mean(sensitivity5), mean(sensitivity1), mean(sensitivity2), mean(sensitivity3), mean(sensitivity4))
    mylist2[[n]] <- c(mean(specificity5), mean(specificity1), mean(specificity2), mean(specificity3), mean(specificity4))
  }

  outdt <- do.call(cbind, mylist1)
  rownames(outdt) <- methods
  colnames(outdt) <- sample_sizes
  x <- 1:dim(outdt)[2]

  par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))
  plot(x, ylim = c(0, 0.6), col = "white", ylab = "Power",
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = "n",
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title(paste("betaf0 =", i1, ", para = (0, 2, 0, 1)"))
  axis(2, at = seq(0, 0.6, 0.1), lab = paste0(seq(0, 60, 10), "%"), las = 2, cex.axis = 1.3)
  axis(1, at = x, lab = colnames(outdt), cex.axis = 1.3)
  grid()
  for (r in 1:5) {
    points(x, outdt[r, ], col = mycol[r], pch = 19, cex = 1.8)
    lines(x, outdt[r, ], col = mycol[r], lwd = 1.5)
  }
  legend("topleft", legend = methods, col = mycol[1:5], lwd = 1, bty = "n",
         lty = rep(NA, 5), pch = rep(19, 5))

  outdt2 <- do.call(cbind, mylist2)
  rownames(outdt2) <- methods
  colnames(outdt2) <- sample_sizes
  plot(x, ylim = c(0, 1), col = "white", ylab = "FPR",
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = "n",
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title("para = (0, 2, 0, 0), FDR = 0.1")
  axis(2, at = seq(0, 1, 0.1), lab = paste0(seq(0, 100, 10), "%"), las = 2, cex.axis = 1.3)
  axis(1, at = x, lab = colnames(outdt2), cex.axis = 1.3)
  grid()
  for (r in 1:5) {
    points(x, outdt2[r, ], col = mycol[r], pch = 19, cex = 1.8)
    lines(x, outdt2[r, ], col = mycol[r], lwd = 1.5)
  }
}
```

### betaf0 = 1

```{r og-conti-f1}
Nsim <- 50
rr <- 0.1
methods <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
sample_sizes <- c(200, 400, 600, 800, 1000, 1200)
mycol <- c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61")
mylist1 <- mylist2 <- list()
m.pvalue <- lapply(1:6, function(x) matrix(nrow = 200, ncol = Nsim))

for (i1 in c(1)) {
  n <- 0
  for (i3 in sample_sizes) {
    n <- n + 1
    sensitivity1=c(); specificity1=c()
    sensitivity2=c(); specificity2=c()
    sensitivity3=c(); specificity3=c()
    sensitivity4=c(); specificity4=c()
    sensitivity5=c(); specificity5=c()
    sensitivity6=c(); specificity6=c()

    for (i2 in 1:200) {
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene/diff/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/Volumes/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene/other/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))

      m.pvalue[[1]][i2,] <- simuresother[["m1.pvalue"]]
      m.pvalue[[2]][i2,] <- simuresother[["m2.pvalue"]]
      m.pvalue[[3]][i2,] <- simuresother[["m3.pvalue"]]
      m.pvalue[[4]][i2,] <- simuresother[["m4.pvalue"]]
      m.pvalue[[5]][i2,] <- simuresdiffFix[["m1.pvalue"]]
      m.pvalue[[6]][i2,] <- simuresdiff[["m1.pvalue"]]
    }
    for (j2 in 1:Nsim) {
      for (k in 1:5) {
        idx <- k
        if (k == 1) idx <- 5
        if (k == 2) idx <- 1
        if (k == 3) idx <- 2
        if (k == 4) idx <- 3
        if (k == 5) idx <- 4
        index1 <- which(p.adjust(m.pvalue[[idx]][, j2], method = "BH") < rr)
        sens <- ifelse(length(index1) != 0, length(which(index1 < 51)) / 50, 0)
        spec <- ifelse(length(index1) != 0, length(which(index1 > 50)) / length(index1), 0)
        if (k == 1) { sensitivity5 <- c(sensitivity5, sens); specificity5 <- c(specificity5, spec) }
        if (k == 2) { sensitivity1 <- c(sensitivity1, sens); specificity1 <- c(specificity1, spec) }
        if (k == 3) { sensitivity2 <- c(sensitivity2, sens); specificity2 <- c(specificity2, spec) }
        if (k == 4) { sensitivity3 <- c(sensitivity3, sens); specificity3 <- c(specificity3, spec) }
        if (k == 5) { sensitivity4 <- c(sensitivity4, sens); specificity4 <- c(specificity4, spec) }
      }
    }
    mylist1[[n]] <- c(mean(sensitivity5), mean(sensitivity1), mean(sensitivity2), mean(sensitivity3), mean(sensitivity4))
    mylist2[[n]] <- c(mean(specificity5), mean(specificity1), mean(specificity2), mean(specificity3), mean(specificity4))
  }

  outdt <- do.call(cbind, mylist1)
  rownames(outdt) <- methods
  colnames(outdt) <- sample_sizes
  x <- 1:dim(outdt)[2]

  par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))
  plot(x, ylim = c(0, 0.6), col = "white", ylab = "Power",
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = "n",
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title(paste("betaf0 =", i1, ", para = (0, 2, 0, 2)"))
  axis(2, at = seq(0, 1, 0.1), lab = paste0(seq(0, 100, 10), "%"), las = 2, cex.axis = 1.3)
  axis(1, at = x, lab = colnames(outdt), cex.axis = 1.3)
  grid()
  for (r in 1:5) {
    points(x, outdt[r, ], col = mycol[r], pch = 19, cex = 1.8)
    lines(x, outdt[r, ], col = mycol[r], lwd = 1.5)
  }

  outdt2 <- do.call(cbind, mylist2)
  rownames(outdt2) <- methods
  colnames(outdt2) <- sample_sizes
  plot(x, ylim = c(0, 1), col = "white", ylab = "FPR",
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = "n",
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title("para = (0, 2, 0, 0), FDR = 0.1")
  axis(2, at = seq(0, 1, 0.1), lab = paste0(seq(0, 100, 10), "%"), las = 2, cex.axis = 1.3)
  axis(1, at = x, lab = colnames(outdt2), cex.axis = 1.3)
  grid()
  for (r in 1:5) {
    points(x, outdt2[r, ], col = mycol[r], pch = 19, cex = 1.8)
    lines(x, outdt2[r, ], col = mycol[r], lwd = 1.5)
  }
}
```
