---
title: "Simulation: Confounding Between Phenotype and Mutational Signatures"
author: "Jie Zhou, Siming Zhao"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: hide
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This simulation demonstrates that when the phenotype is correlated with
mutational signatures — but **not** with selection — benchmark methods produce
false positives while DiffDriver correctly controls the false positive rate.

### Simulation setup

- **Background mutation rate (realistic, LUAD-derived)**: The per-site
  background probability is `bmr = (signatures %*% loadings) / sc`, built entirely
  from LUAD data — so it is *not* an arbitrary rate:
    - **Shape**: the `signatures` (96 × 6) and per-sample `loadings` (`ll$LUAD`)
      are the real LUAD mutational signatures and exposures (fastTopics on the
      LUAD mutation counts, 527 samples, bootstrapped up to 1,500). The loadings
      are a composition (constant column sums = 62.75), so a summed-loading
      covariate carries no phenotype information and cannot adjust the confounding
      away.
    - **Level**: `sc = 359000` gives a mean per-site probability of ≈1.8×10⁻⁶ and
      a genome-wide **synonymous burden of ≈63 mutations per sample** (median 62,
      range 56–73), matching the **observed ≈64 synonymous mutations per sample**
      in real TCGA-LUAD (530 samples). So both the spectrum and the overall level
      are LUAD-realistic.
- **Mutational signatures**: Six signatures extracted from LUAD using
  [fastTopics](https://stephenslab.github.io/fastTopics/). Signature 3 — a
  T>G–dominated (smoking-like) signature whose highest-loading context is
  `C[T>G]G` (nucleotide type 34) — is selected as the confounding signature.
- **Phenotype**: Simulated to correlate with signature 3 ($R^2 = 0.78$), but
  with **zero correlation** with selection (`para = c(0.5, 0.5)`).
- **Mutations**: Simulated for a single gene (ERBB3) using oncogene mutation
  rates (background + selection), with two hotspots placed at nucleotide type 34
  (the context with the highest loading in signature 3), so that the
  hotspot-dominated mutations track the phenotype.

### Key result

Because the phenotype correlates with a mutational signature that affects
overall mutation rates, the count-based benchmark methods (Linear regression,
Fisher's exact test, Binomial test, Logistic regression, and the Mann-Whitney
rank-sum test) increasingly detect differential selection as sample size grows —
all false positives. DiffDriver, which accounts for signature-based background
mutation rates, maintains a low false positive rate.

[coselens](https://github.com/ggonzalezlopez/coselens) is also included as a
"true-background" reference: the phenotype is binarized into two groups and
coselens is given each group's **real** mutation rate (the group-summed neutral
rate per trinucleotide context from the simulation, assuming the rate is the same
across samples within a group; see [coselens setup](#coselens-with-the-true-per-group-rate)).
Because it is handed the true confounded background, coselens correctly
attributes the case/control mutation excess to background rather than selection
and also controls the false positive rate.

## Rerunning the simulation

The simulation functions live in [`code/simulate_functions.R`](https://github.com/szhaolab/diffdriver-paper/blob/main/code/simulate_functions.R)
under the "CONFOUNDING scenario" section (`build_confounded_bmr`,
`simulate_confounding`, `run_confounding_benchmarks`,
`run_confounding_diffdriver`, `run_confounding_diffdriver_fixed`). The benchmark
methods are all in `run_confounding_benchmarks`
(Linear/Fisher/Binomial/LogisticR + Mann-Whitney); DiffDriver is run by
`run_confounding_diffdriver_fixed`.

The inputs are bundled in `data/bmr_sim_inputs.rda` (`beta_gc`, `beta_gcFix`,
`signature`, `ll`, `sganno`, `sgmatrix`), together with `data/hmm.rda`,
`data/codeSignature.rda`, and `data/hotspot1sig.rda` (hotspots at nttype 34).
A single sample size is reproduced with:

```{r eval=FALSE}
library(diffdriver); library(data.table); library(Matrix)
source("code/simulate_functions.R")
load("data/bmr_sim_inputs.rda"); load("data/hmm.rda")
load("data/codeSignature.rda");  load("data/hotspot1sig.rda")

set.seed(1)
res_other <- run_confounding_benchmarks(
  binary = TRUE, Niter = 100,
  sganno = bmr_sim$sganno, sgmatrix = bmr_sim$sgmatrix,
  Nsample = 1500, para = c(0.5, 0.5),      # no differential selection
  rho = sqrt(0.78),                        # phenotype ~ signature 3, R^2 = 0.78
  signatures = bmr_sim$signature, loadings = bmr_sim$ll,
  beta_gc = bmr_sim$beta_gc, hot = 1, hmm = hmm,
  sc = 359000)                             # background-rate scale (~188 mutations)

# false-positive rate per method (p < 0.01), no true signal so all are FPs
sapply(res_other[c("m1.pvalue","m2.pvalue","m3.pvalue","m4.pvalue",
                    "mannwhitney.pvalue")], function(p) mean(p < 0.01))
```

The full benchmark loops this over `Nsample = 200 … 1500`; the Mann-Whitney
p-values used below were produced this way and saved to
`data/pppower_mannwhitney_bmr.Rd`. Because the original
`scripts/simulate_functions.R` (and the exact HMM coefficients) were lost, the
rerun is *close* to — but not bit-identical to — the archived runs; the Linear
and Fisher/Binomial/LogisticR curves plotted below are read from the original
archived results.

### coselens with the true per-group rate

coselens compares dN/dS between two groups. We binarize the phenotype into cases
(E = 1) and controls (E = 0) and, rather than letting dndscv estimate the
background, feed each group its **true** neutral rate from the simulation. Under
the assumption that the rate is the same across samples within a group, a group's
per-context rate is the group-summed neutral rate `rowSums(bmrfold$bmr[, group])`;
this 96-context vector is mapped to dndscv's 192 substitution contexts via
`data/context96_map.rds` and passed to `dndscv_truerate` (a rate-fixed dndscv).
`code/run_coselens.R` wraps this; the per-iteration driver is:

```{r eval=FALSE}
source("code/run_coselens.R"); source("code/dndscv_truerate.R")
library(coselens)
context96_map <- readRDS("data/context96_map.rds")

# `sim` is one cohort returned by simulate_confounding() (same set.seed(1) loop
# as above, so the cohorts are identical to the other methods).
e   <- sim$pheno; g1 <- which(e == 1); g2 <- which(e == 0)
maf <- sim_to_maf(sim$mutlist, sim$annodata, e)           # case/control MAFs
mr1 <- rowSums(sim$bmrfold$bmr[, g1])[context96_map$number]  # 192-context true rate
mr2 <- rowSums(sim$bmrfold$bmr[, g2])[context96_map$number]
res <- run_coselens(maf$group1, maf$group2, subset.genes.by = "ERBB3",
                    true_rate = list(mutrates1 = mr1, mutrates2 = mr2, theta = 1e6),
                    gene_list = "ERBB3", cv = NULL)
res$substitutions$pval          # differential-selection p-value for ERBB3
```

coselens was run on the **same simulated cohorts** as the other methods (verified:
per-iteration mutation counts are identical across all 100 iterations and 7 sample
sizes). Its p-values are saved to `data/pppower_coselens_bmr.Rd`.



```{r}
Nsim=100
SIZES <- c(200,400,600,800,1000,1200,1500)

# Mann-Whitney and coselens FPs from the rerun on the SAME simulated cohorts
# (code/simulate_functions.R::run_confounding_benchmarks + code/run_coselens.R).
load("data/pppower_mannwhitney_bmr.Rd")   # object `mw`,       keyed by sample size
load("data/pppower_coselens_bmr.Rd")      # object `coselens`, keyed by sample size

mylist <- list()
for (i1 in c(1)){
   for (rho in c(0)) {
    for (i4 in c(0.5)) {
    n<-0
    for(i3 in SIZES){
      n <- n+1
      load(paste0("data/bmr_archive/diff/pppower_betaf0=",i1, "_sample",i3,"_",i4,"_",rho,".Rd"))
          load(paste0("data/bmr_archive/other/pppower_betaf0=",i1,"_sample",i3,"_",i4,"_",rho,".Rd"))

      m5.pvalue <- simuresdiff[["m1.pvalue"]]
      m1.pvalue <- simuresother[["m1.pvalue"]]
      m2.pvalue <- simuresother[["m2.pvalue"]]
      m3.pvalue <- simuresother[["m3.pvalue"]]
      m4.pvalue <- simuresother[["m4.pvalue"]]
      m7.pvalue <- mw[[as.character(i3)]][["mannwhitney.pvalue"]]       # Mann-Whitney
      m8.pvalue <- coselens[[as.character(i3)]][["coselens.pvalue"]]    # coselens (true per-group BMR)
     mylist[[n]] <- c(length(m1.pvalue[m1.pvalue <0.01]),length(m2.pvalue[m2.pvalue <0.01]),length(m3.pvalue[m3.pvalue <0.01]), length(m4.pvalue[m4.pvalue <0.01]),length(m7.pvalue[m7.pvalue <0.01]),length(m8.pvalue[m8.pvalue <0.01]),length(m5.pvalue[m5.pvalue <0.01]))/Nsim
    }
    outdt <- do.call(cbind, mylist)
    rownames(outdt)=c("Linear","Fisher","Binomial","LogisticR","MannWhitney","Coselens","diffDriver")
    colnames(outdt) <- SIZES
    # colors: Linear, Fisher, Binomial, LogisticR, MannWhitney, coselens, diff
    mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#1a9850","#ff7f00","#FF3D2E")
    mypch=c(19,19,19,19,3,8,17)
     plot(1:dim(outdt)[2],ylim=c(0,1), col="white", ylab = "False positive rate (p < 0.01)", xlab="Number of samples", yaxt="n", xaxt="n")
     title( paste( "SC(scale efficient)=",i1,",","para=(",i4,1-i4,")", ",", sep=" "))
    yticks_val <- seq(0,1,0.1)
    axis(2, at=yticks_val, lab=paste0(yticks_val),las=2)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt))
    grid()
    for (r in 1:7){
      points(outdt[r,],col=mycol[r], pch=mypch[r], cex=1.3)
      lines(outdt[r,],col=mycol[r])
    }
    legend("topleft", legend = c("Linear","Fisher","Binomial","LogisticR","MannWhitney","Coselens","diff"), col= mycol, lwd=1, bty="n", lty=rep(NA,7), pch=mypch)
  }
}
  }

```
