---
title: "Simulation: Pipeline and Power Results"
author: "Jie Zhou, Siming Zhao"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
---

## Overview

This document runs a simulation study to evaluate the power of the DiffDriver
model. The simulation pipeline has three steps:

1. **Simulate phenotype and mutations** — Given a gene's positional annotation
   and functional features, generate binary (case/control) phenotypes and
   somatic mutations under a selection model (`simulate_selection`).
2. **Test for differential driver signal** — Apply `ddmodel_binary` (or
   `ddmodel` for continuous phenotypes) to the simulated data to obtain a
   p-value (`run_one_iteration`).
3. **Repeat** — Run multiple iterations to estimate power (`run_simulation`).

**Note:** The [Run simulation with a single gene](#run-simulation-with-a-single-gene)
section below demonstrates the pipeline for one gene. The
[Run simulation on 200 genes](#run-simulation-on-200-genes) section then reports the
full aggregated comparison, reproducing Figure 3 in the paper.

### Source files

| File | Description |
|------|-------------|
| `code/simulate_functions.R` | `hotspotseq()` — generates hotspot indicator sequence via HMM; `simulate_selection()` — simulates phenotypes, selection, and mutations for one gene (background rate is constant across samples); `power_comparediffiFix()` / `power_compareotheri()` — run `simulate_selection()` over `Niter` iterations and apply DiffDriver / the benchmark tests |
| `code/run_simulation.R`     | `run_one_iteration()` — single iteration: simulate then test; `run_simulation()` — loop over iterations |
| `code/simulate_all_genes.R` | master script: loops `power_compare*()` over all 200 genes for a given setting (see [Data generation](#data-generation)) |

## Setup

```{r setup, message=FALSE}
library("diffdriver")
library("data.table")
source("code/simulate_functions.R")
source("code/run_simulation.R")
set.seed(1)
```

### Load data

- **ri200** — list of 239 genes; each element is a list of 9 annotation
  data.frames (one per nucleotide type), containing positional and mutation
  type information.
- **fanno200** — matching list of functional annotation matrices (features
  used in the selection model: intercept, functypecode8, conservation scores,
  etc.).
- **beta_gc** — TSG functional effect coefficients (from `TSGpars`):
  intercept, functypecode8, mycons, sift, phylop100, MA.
- **BMR** — background mutation rates, one per nucleotide type (9 values).
- **hmm** — HMM parameters for hotspot detection: transition probabilities
  and logistic regression coefficients (10 values).

```{r load-data}
load("data/ri200.Rd")
load("data/fanno200.Rd")
load("data/beta_tsg.Rd")   # loads beta_gc
load("data/BMR.rda")        # loads BMR
load("data/hmm.rda")        # loads hmm
```

## Simulation parameters

| Parameter | Value | Description |
|-----------|-------|-------------|
| `betaf0`  | 1     | Baseline functional effect size (log-scale fold change) |
| `number_sample` | 1200 | Total number of samples (split equally into case/control) |
| `gene_index` | 1  | Index into ri200/fanno200 for the target gene |
| `ifbinary` | TRUE | Use binary (case/control) phenotype |
| `Niter`   | 1     | Number of simulation iterations |
| `para`    | (0.5, 0.5) | Binary: c(selection prob for cases, selection prob for controls). Continuous: c(mean, sd, intercept, slope) where intercept and slope are logistic regression coefficients determining selection probability as a function of phenotype |
| `hot`     | 0     | Hotspot flag (0 = no hotspot effect) |

```{r parameters}
betaf0 <- 1
number_sample <- 1200
gene_index <- 1 # pick a gene from the 239 genes to run.
ifbinary <- TRUE
Niter <- 1
para <- c(0.5, 0.5)
```

## Run simulation with a single gene

```{r run-simulation}
simuresdiff <- run_simulation(
  binary   = ifbinary,
  Niter    = Niter,
  sganno   = ri200[[gene_index]],
  sgmatrix = fanno200[[gene_index]],
  bmrpars  = log(BMR),
  Nsample  = number_sample,
  betaf0   = betaf0,
  beta_gc  = beta_gc,
  para     = para,
  hot      = 0,
  hmm      = hmm
)
```

```{r results}
cat("P-values:", simuresdiff$pvalues, "\n")
cat("Number of mutations:", simuresdiff$nummuts, "\n")
```

---


## Run simulation on 200 genes

For 200 genes with five functional annotations, we compare the power (sensitivity)
and false positive rate (FPR) of DiffDriver against benchmark methods: Linear
regression, Fisher's exact test, Binomial test, and Logistic Regression.

The first 50 genes have differential selection while the remaining 150 do not.
FDR is controlled at 0.1 using the BH procedure.

### Simulation parameters

These simulations use the same pipeline described in
the [simulation pipeline](#overview) above. The key parameters are:

| Parameter | Description |
|-----------|-------------|
| `betaf0`  | Baseline functional effect size: 0 or 1 |
| `para`    | Binary: selection probabilities c(prob_case, prob_control). Continuous: c(mean, sd, intercept, slope) |
| `beta_gc` | Simulation functional effect coefficients — the true effect sizes used to generate mutations (from `data/beta_tsg.Rd` or `data/beta_og.Rd`) |
| `Nsample` | Varies: 200, 400, 600, 800, 1000, 1200, 1500 |
| `Niter`   | 100 (binary) or 50 (continuous) per gene |

### Functional effect coefficients

Two distinct coefficient sets are involved and should not be confused.

**Simulation coefficients** (`beta_gc`) — the *true* functional effect sizes used
to generate the mutations, loaded from `data/beta_tsg.Rd` / `data/beta_og.Rd`:

- **TSG**: `(Intercept)=0.851, functypecode8=0.614, mycons=0.375, sift=0.166, phylop100=0.358, MA=0.223`
- **OG**:  `(Intercept)=0.450, functypecode8=0.049, mycons=0.806, sift=-0.078, phylop100=0.664, MA=0.089`

**DiffDriver built-in parameters** — the fixed coefficients DiffDriver uses when it
is *run* (the `diffFix` method's `beta_gcFix`). These are a property of the fitted
model, not of the simulation:

- **TSG**: `(Intercept)=0.851, functypecode8=0.458, mycons=0.137, sift=0.124, phylop100=0.083, MA=0.306`
- **OG**:  `(Intercept)=0.450, functypecode8=-0.186, mycons=0.316, sift=0.020, phylop100=0.256, MA=0.140`

### Data generation

All result files below were produced by the master script
`code/simulate_all_genes.R`, which loops over all 200 genes for a given setting
(consolidating the original per-gene scripts). It is run once per **setting** and
**method**:

```bash
Rscript code/simulate_all_genes.R <SETTING> <METHOD>
```

`METHOD` is `diffFix` (DiffDriver) or `other` (the benchmark methods: Linear,
Fisher, Binomial, Logistic); each folder stores the two methods' `.Rd` outputs in
its `diffFix/` and `other/` subdirectories. `SETTING` selects the gene set,
phenotype, and selection probability. The exact command for each folder is shown
with its results below.

The master script sources `code/simulate_functions.R`, so the mutations are
generated by the same `simulate_selection()` function documented in the
[pipeline](#overview) above (the `power_comparediffiFix` / `power_compareotheri`
drivers there call it and apply DiffDriver / the benchmark tests); the diffdriver
package supplies the underlying model and test functions.

---

```{r setup-power, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

METHODS      <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
# Colours 6+ are for the extra methods (coselens, Mann-Whitney), added only for
# settings that have the corresponding pppower_all_<SETTING>_<method>.Rd (load_and_compute).
MYCOL        <- c("#FF3D2E", "#00AFBB", "#aa00ff", "#0001f6", "#fdae61", "#1a9850", "#a65628")
FDR          <- 0.1
BINARY_SIZES <- c(200, 400, 600, 800, 1000, 1200, 1500)
CONTI_SIZES  <- c(200, 400, 600, 800, 1000, 1200)

# Each setting's consolidated results (pppower_all_<SETTING>_<METHOD>.Rd, produced by
# code/consolidate_results.R) live in that setting's original folder under DD.
DD <- "/Volumes/Szhao/jiezhou/diffdata"
SETTING_DIR <- c(
  TSG_binary_0.8 = file.path(DD, "gene239/conti/TSG_BRCA"),
  TSG_binary_0.9 = file.path(DD, "gene239/conti/TSG_BRCA/TSG0.9/TSG_gene"),
  TSG_conti      = file.path(DD, "gene239/conti/TSG_BRCA/conti/conti2"),
  OG_binary_0.8  = file.path(DD, "binaryEM/OG_gene"),
  OG_binary_0.9  = file.path(DD, "binaryEM/OG_gene"),
  OG_conti       = file.path(DD, "gene239/conti/OCG/OG_gene")
)

#' Compute power and FPR from p-value matrices across simulation iterations.
#' @param m.pvalue List of matrices (one per method), each n_genes x Nsim.
#' @param Nsim Number of simulation iterations.
#' @param n_true Number of true positive genes (first n_true genes).
compute_metrics <- function(m.pvalue, Nsim, n_true = 50) {
  n_methods <- length(m.pvalue)
  power <- fpr <- numeric(n_methods)
  for (k in seq_len(n_methods)) {
    sens <- fpr_vals <- numeric(Nsim)
    for (j in seq_len(Nsim)) {
      hits <- which(p.adjust(m.pvalue[[k]][, j], method = "BH") < FDR)
      if (length(hits) > 0) {
        sens[j]     <- sum(hits <= n_true) / n_true
        fpr_vals[j] <- sum(hits > n_true) / length(hits)
      }
    }
    power[k] <- mean(sens)
    fpr[k]   <- mean(fpr_vals)
  }
  list(power = power, fpr = fpr)
}

#' Load consolidated simulation results and compute power/FPR per sample size.
#'
#' Each setting has two consolidated files in its folder (SETTING_DIR[[setting]]),
#' pppower_all_<setting>_diffFix.Rd and _other.Rd, each holding a list
#' `simures_all` keyed by the per-gene filename stem. The p-values are:
#'   DiffDriver                       = diffFix result$m1.pvalue
#'   Linear/Fisher/Binomial/LogisticR = other   result$m1--m4.pvalue
#'
#' @param setting Setting name (e.g. "TSG_binary_0.8") selecting the two files.
#' @param file_fn Function(Nsample, gene_index) returning the per-gene .Rd name;
#'   its stem (without .Rd) is the key into simures_all.
#' @param sample_sizes Vector of sample sizes to evaluate.
#' @param Nsim Max number of iterations to use. If NULL, uses all available.
#' @param n_genes Number of genes (default 200).
load_and_compute <- function(setting, file_fn, sample_sizes,
                             Nsim = NULL, n_genes = 200) {
  base <- SETTING_DIR[[setting]]
  ed <- new.env(); load(file.path(base, sprintf("pppower_all_%s_diffFix.Rd", setting)), envir = ed)
  eo <- new.env(); load(file.path(base, sprintf("pppower_all_%s_other.Rd",   setting)), envir = eo)
  diff_all  <- ed$simures_all
  other_all <- eo$simures_all

  # Extra methods (coselens, Mann-Whitney) are added only where a consolidated file
  # exists for the setting; each gene key holds a <field> p-value vector like the others.
  EXTRA     <- list(coselens = "coselens.pvalue", MannWhitney = "mannwhitney.pvalue")
  EXTRA_SFX <- c(coselens = "coselens", MannWhitney = "mannwhitney")
  extra_all <- list()
  for (nm in names(EXTRA)) {
    f <- file.path(base, sprintf("pppower_all_%s_%s.Rd", setting, EXTRA_SFX[[nm]]))
    if (file.exists(f)) { ee <- new.env(); load(f, envir = ee); extra_all[[nm]] <- ee$simures_all }
  }
  methods   <- c(METHODS, names(extra_all))
  n_methods <- length(methods)

  power_mat <- fpr_mat <- matrix(
    nrow = n_methods, ncol = length(sample_sizes),
    dimnames = list(methods, sample_sizes)
  )
  for (s in seq_along(sample_sizes)) {
    N <- sample_sizes[s]
    m.pvalue <- NULL
    n_iter   <- NULL
    for (g in seq_len(n_genes)) {
      key   <- sub("\\.Rd$", "", file_fn(N, g))
      sdiff <- diff_all[[key]]
      soth  <- other_all[[key]]
      if (is.null(n_iter)) {
        n_avail <- length(sdiff[["m1.pvalue"]])
        n_iter  <- if (!is.null(Nsim)) min(Nsim, n_avail) else n_avail
        m.pvalue <- lapply(seq_len(n_methods), function(x)
          matrix(nrow = n_genes, ncol = n_iter))
      }
      idx <- seq_len(n_iter)
      m.pvalue[[1]][g, ] <- sdiff[["m1.pvalue"]][idx]
      m.pvalue[[2]][g, ] <- soth[["m1.pvalue"]][idx]
      m.pvalue[[3]][g, ] <- soth[["m2.pvalue"]][idx]
      m.pvalue[[4]][g, ] <- soth[["m3.pvalue"]][idx]
      m.pvalue[[5]][g, ] <- soth[["m4.pvalue"]][idx]
      for (ei in seq_along(extra_all)) {
        nm <- names(extra_all)[ei]
        m.pvalue[[length(METHODS) + ei]][g, ] <- extra_all[[nm]][[key]][[EXTRA[[nm]]]][idx]
      }
    }
    metrics <- compute_metrics(m.pvalue, n_iter)
    power_mat[, s] <- metrics$power
    fpr_mat[, s]   <- metrics$fpr
  }
  list(power = power_mat, fpr = fpr_mat, methods = methods)
}

#' Plot power and FPR side by side.
#' @param res Output from load_and_compute (list with power and fpr matrices).
#' @param power_title Title for the power panel.
#' @param fpr_title Title for the FPR panel.
#' @param power_ylim Y-axis upper limit for the power panel.
#' @param show_legend Whether to show the method legend on the power panel.
plot_power_fpr <- function(res, power_title, fpr_title,
                           power_ylim = 0.4, show_legend = TRUE) {
  x     <- seq_len(ncol(res$power))
  sizes <- colnames(res$power)
  par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))

  # Power panel
  plot(x, ylim = c(0, power_ylim), col = "white", ylab = "Power",
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = "n",
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title(power_title)
  yticks <- seq(0, power_ylim, 0.1)
  axis(2, at = yticks, lab = paste0(yticks * 100, "%"), las = 2, cex.axis = 1.3)
  axis(1, at = x, lab = sizes, cex.axis = 1.3)
  grid()
  for (r in seq_len(nrow(res$power))) {
    points(x, res$power[r, ], col = MYCOL[r], pch = 19, cex = 1.8)
    lines(x, res$power[r, ], col = MYCOL[r], lwd = 1.5)
  }
  if (show_legend) {
    mm <- rownames(res$power)
    legend("topleft", legend = mm, col = MYCOL[seq_along(mm)],
           lwd = 1, bty = "n", lty = NA, pch = 19)
  }

  # FPR panel
  plot(x, ylim = c(0, 1), col = "white", ylab = "FPR",
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty = "n",
       cex.lab = 1.3, mgp = c(4, 1, 0))
  title(fpr_title)
  axis(2, at = seq(0, 1, 0.1), lab = paste0(seq(0, 100, 10), "%"),
       las = 2, cex.axis = 1.3)
  axis(1, at = x, lab = sizes, cex.axis = 1.3)
  grid()
  for (r in seq_len(nrow(res$fpr))) {
    points(x, res$fpr[r, ], col = MYCOL[r], pch = 19, cex = 1.8)
    lines(x, res$fpr[r, ], col = MYCOL[r], lwd = 1.5)
  }
}
```

## Reproducing Figure 3 in paper

The panels below reproduce Figure 3: power (sensitivity) and false positive rate
(FPR) versus sample size for DiffDriver against the benchmark methods, with one
setting per subsection.

### TSG, Binary Phenotype

#### para = c(0.8, 0.2)

**Generated by:**

```bash
Rscript code/simulate_all_genes.R TSG_binary_0.8 diffFix
Rscript code/simulate_all_genes.R TSG_binary_0.8 other
```

```{r tsg-binary-08, fig.width=12, fig.height=5}
res <- load_and_compute(
  setting      = "TSG_binary_0.8",
  file_fn      = function(N, g) paste0("pppower_betaf0=0_sample", N, "_", g, ".Rd"),
  sample_sizes = BINARY_SIZES
)
plot_power_fpr(res,
  power_title = "para = (0.8, 0.2)",
  fpr_title   = "para = (0.5, 0.5), FDR = 0.1"
)
```

#### para = c(0.9, 0.1)

**Generated by:**

```bash
Rscript code/simulate_all_genes.R TSG_binary_0.9 diffFix
Rscript code/simulate_all_genes.R TSG_binary_0.9 other
```

```{r tsg-binary-09, fig.width=12, fig.height=5}
res <- load_and_compute(
  setting      = "TSG_binary_0.9",
  file_fn      = function(N, g) paste0("pppower_betaf0=0_sample", N, "_", g, ".Rd"),
  sample_sizes = BINARY_SIZES
)
plot_power_fpr(res,
  power_title  = "para = (0.9, 0.1)",
  fpr_title    = "para = (0.5, 0.5), FDR = 0.1",
  show_legend  = FALSE
)
```

---

### TSG, Continuous Phenotype

Continuous phenotype with `para = c(0, 2, 0, 1)`: phenotype ~ N(0, 2), selection
probability determined by logistic regression with intercept = 0 and slope = 1.

**Generated by** (one run produces both betaf0 = 0 and betaf0 = 1):

```bash
Rscript code/simulate_all_genes.R TSG_conti diffFix
Rscript code/simulate_all_genes.R TSG_conti other
```

#### betaf0 = 0

```{r tsg-conti-f0, fig.width=12, fig.height=5}
res <- load_and_compute(
  setting      = "TSG_conti",
  file_fn      = function(N, g) paste0("pppower_betaf0=0_sample", N, "_", g, ".Rd"),
  sample_sizes = CONTI_SIZES
)
plot_power_fpr(res,
  power_title = "betaf0 = 0, para = (0, 2, 0, 1)",
  fpr_title   = "para = (0, 2, 0, 0), FDR = 0.1",
  power_ylim  = 0.6
)
```

#### betaf0 = 1

```{r tsg-conti-f1, fig.width=12, fig.height=5}
res <- load_and_compute(
  setting      = "TSG_conti",
  file_fn      = function(N, g) paste0("pppower_betaf0=1_sample", N, "_", g, ".Rd"),
  sample_sizes = CONTI_SIZES
)
plot_power_fpr(res,
  power_title  = "betaf0 = 1, para = (0, 2, 0, 2)",
  fpr_title    = "para = (0, 2, 0, 0), FDR = 0.1",
  power_ylim   = 0.6,
  show_legend  = FALSE
)
```

---

### OG, Binary Phenotype

#### para = c(0.8, 0.2)

**Generated by:**

```bash
Rscript code/simulate_all_genes.R OG_binary_0.8 diffFix
Rscript code/simulate_all_genes.R OG_binary_0.8 other
```

```{r og-binary-08, fig.width=12, fig.height=5}
res <- load_and_compute(
  setting      = "OG_binary_0.8",
  file_fn      = function(N, g) paste0("pppower_betaf0=0_sample", N, "_0.8_", g, ".Rd"),
  sample_sizes = BINARY_SIZES
)
plot_power_fpr(res,
  power_title = "para = (0.8, 0.2)",
  fpr_title   = "para = (0.5, 0.5), FDR = 0.1",
  power_ylim  = 0.5
)
```

#### para = c(0.9, 0.1)

**Generated by:**

```bash
Rscript code/simulate_all_genes.R OG_binary_0.9 diffFix
Rscript code/simulate_all_genes.R OG_binary_0.9 other
```

```{r og-binary-09, fig.width=12, fig.height=5}
res <- load_and_compute(
  setting      = "OG_binary_0.9",
  file_fn      = function(N, g) paste0("pppower_betaf0=0_sample", N, "_0.9_", g, ".Rd"),
  sample_sizes = BINARY_SIZES
)
plot_power_fpr(res,
  power_title  = "para = (0.9, 0.1)",
  fpr_title    = "para = (0.5, 0.5), FDR = 0.1",
  power_ylim   = 0.5,
  show_legend  = FALSE
)
```

---

### OG, Continuous Phenotype

Continuous phenotype with `para = c(0, 2, 0, 1)`: phenotype ~ N(0, 2), selection
probability determined by logistic regression with intercept = 0 and slope = 1.

**Generated by** (one run produces both betaf0 = 0 and betaf0 = 1):

```bash
Rscript code/simulate_all_genes.R OG_conti diffFix
Rscript code/simulate_all_genes.R OG_conti other
```

#### betaf0 = 0

```{r og-conti-f0, fig.width=12, fig.height=5}
res <- load_and_compute(
  setting      = "OG_conti",
  file_fn      = function(N, g) paste0("pppower_betaf0=0_sample", N, "_", g, ".Rd"),
  sample_sizes = CONTI_SIZES
)
plot_power_fpr(res,
  power_title = "betaf0 = 0, para = (0, 2, 0, 1)",
  fpr_title   = "para = (0, 2, 0, 0), FDR = 0.1",
  power_ylim  = 0.6
)
```

#### betaf0 = 1

```{r og-conti-f1, fig.width=12, fig.height=5}
res <- load_and_compute(
  setting      = "OG_conti",
  file_fn      = function(N, g) paste0("pppower_betaf0=1_sample", N, "_", g, ".Rd"),
  sample_sizes = CONTI_SIZES
)
plot_power_fpr(res,
  power_title  = "betaf0 = 1, para = (0, 2, 0, 2)",
  fpr_title    = "para = (0, 2, 0, 0), FDR = 0.1",
  power_ylim   = 0.6,
  show_legend  = FALSE
)
```
