---
title: "Simulation Study for DiffDriver"
author: "Jie Zhou, Siming Zhao"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
---

## Overview

This document runs a simulation study to evaluate the power of the DiffDriver
model. The simulation pipeline has three steps:

1. **Simulate phenotype and mutations** — Given a gene's positional annotation
   and functional features, generate binary (case/control) phenotypes and
   somatic mutations under a selection model (`simulate_selection`).
2. **Test for differential driver signal** — Apply `ddmodel_binary` (or
   `ddmodel` for continuous phenotypes) to the simulated data to obtain a
   p-value (`run_one_iteration`).
3. **Repeat** — Run multiple iterations to estimate power (`run_simulation`).

### Source files

| File | Description |
|------|-------------|
| `code/simulate_functions.R` | `hotspotseq()` — generates hotspot indicator sequence via HMM; `simulate_selection()` — simulates phenotypes, selection, and mutations for one gene |
| `code/run_simulation.R`     | `run_one_iteration()` — single iteration: simulate then test; `run_simulation()` — loop over iterations |

## Setup

```{r setup, message=FALSE}
library("diffdriver")
library("data.table")
source("code/simulate_functions.R")
source("code/run_simulation.R")
set.seed(1)
```

### Load data

- **ri200** — list of 239 genes; each element is a list of 9 annotation
  data.frames (one per nucleotide type), containing positional and mutation
  type information.
- **fanno200** — matching list of functional annotation matrices (features
  used in the selection model: intercept, functypecode8, conservation scores,
  etc.).
- **beta_gc** — TSG functional effect coefficients (from `TSGpars`):
  intercept, functypecode8, mycons, sift, phylop100, MA.
- **BMR** — background mutation rates, one per nucleotide type (9 values).
- **hmm** — HMM parameters for hotspot detection: transition probabilities
  and logistic regression coefficients (10 values).

```{r load-data}
load("data/ri200.Rd")
load("data/fanno200.Rd")
load("data/beta_tsg.Rd")   # loads beta_gc
load("data/BMR.rda")        # loads BMR
load("data/hmm.rda")        # loads hmm
```

## Simulation parameters

| Parameter | Value | Description |
|-----------|-------|-------------|
| `betaf0`  | 1     | Baseline functional effect size (log-scale fold change) |
| `number_sample` | 120 | Total number of samples (split equally into case/control) |
| `rho`     | 0.4   | Correlation between phenotype and BMR fold change |
| `gene_index` | 1  | Index into ri200/fanno200 for the target gene |
| `ifbinary` | TRUE | Use binary (case/control) phenotype |
| `Niter`   | 1     | Number of simulation iterations |
| `para`    | (0.5, 0.5) | Binary: c(selection prob for cases, selection prob for controls). Continuous: c(mean, sd, intercept, slope) where intercept and slope are logistic regression coefficients determining selection probability as a function of phenotype |
| `tau`     | 1     | BMR fold-change scaling factor |
| `hot`     | 0     | Hotspot flag (0 = no hotspot effect) |

```{r parameters}
betaf0 <- 1
number_sample <- 1200
rho <- 0
gene_index <- 1 # pick a gene from the 239 genes to run. 
ifbinary <- TRUE
Niter <- 1
para <- c(0.5, 0.5)
```

## Run simulation

```{r run-simulation}
simuresdiff <- run_simulation(
  binary   = ifbinary,
  Niter    = Niter,
  sganno   = ri200[[gene_index]],
  sgmatrix = fanno200[[gene_index]],
  bmrpars  = log(BMR),
  Nsample  = number_sample,
  betaf0   = betaf0,
  beta_gc  = beta_gc,
  para     = para,
  rho      = rho,
  tau      = 1,
  hot      = 0,
  hmm      = hmm
)
```

## Results

```{r results}
cat("P-values:", simuresdiff$pvalues, "\n")
cat("Number of mutations:", simuresdiff$nummuts, "\n")
```
