---
title: "Selected Simulation results"
author: "Jie Zhou, Siming Zhao"
date: "`r Sys.Date()`"
output: 
  workflowr::wflow_html:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: hide
---


```{r}
knitr::opts_chunk$set(echo = TRUE)
```


## Results 2: for TSG and binary phenotype, diffdriver has high power while low FDR


For 200 genes, using five functional annotations, we compare the sensitivity and false discovery rate of different algorithms using binary phenotype based on robust brglm algorithm.  The first 50 genes have differential selection while the left 150 genes do not. The parameter are
`beta_f0=0,1`, `beta_gc=(0.8,0.2)`.For functional effects,old TSG is  drivermap's TSG parameter for   `(functypecode8,       mycons,         sift,    phylop100,           MA, intercept)= (0.45817859, 0.13676461, 0.12436937, 0.08272142, 0.30626015, 0.85127726)`  , while new BRCA's TSG parameter is `((Intercept), functypecode8,        mycons,          sift,     phylop100)=(0.8512773,     0.6136798,     0.3753932,     0.1661389,     0.358036,    0.2229025)` .  The diffdrdiver model use new TSG parameter to simulate and infer while diffdriverFix use new parameter to simulate data and old parameter to select genes, diffdriverFix2 use old parameter to simulate data and infer parameter. 




### para=c(0.8,0.2)

```{r}
Nsim <- 100
rr <- 0.1
mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61","#FFFF00", "#2b2930")
methods <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
sample_sizes <- c(200, 400, 600, 800, 1000, 1200, 1500)
mylist1 <- mylist2 <- replicate(length(sample_sizes), replicate(length(methods), c(), simplify = FALSE), simplify = FALSE)
m.pvalue <- lapply(1:length(methods), function(x) matrix(nrow = 200, ncol = Nsim))

for (i1 in 0) {
  n <- 0
  for (i3 in sample_sizes) {
    n <- n + 1
    for (i2 in 1:200) {
      load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/diff/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/other/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
      load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/diffFixFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))

      m.pvalue[[1]][i2, ] <- simuresdiffFix[["m1.pvalue"]]
      m.pvalue[[2]][i2, ] <- simuresother[["m1.pvalue"]]
      m.pvalue[[3]][i2, ] <- simuresother[["m2.pvalue"]]
      m.pvalue[[4]][i2, ] <- simuresother[["m3.pvalue"]]
      m.pvalue[[5]][i2, ] <- simuresother[["m4.pvalue"]]
    }
    for (j2 in 1:Nsim) {
      for (k in 1:length(methods)) {
        index1 <- which(p.adjust(m.pvalue[[k]][, j2], method = "BH") < rr)
        sensitivity <- ifelse(length(index1) != 0, length(which(index1 < 51)) / 50, 0)
        specificity <- ifelse(length(index1) != 0, length(which(index1 > 50)) / length(index1), 0)
        mylist1[[n]][[k]] <- c(sensitivity, mylist1[[n]][[k]])
        mylist2[[n]][[k]] <- c(specificity, mylist2[[n]][[k]])
      }
    }
  }
}

# pdf("/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/paper/2.a.pdf", width = 7.8, height = 5.5)
par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))
for (i in 1:2) {
  if (i == 1){
    mylist <- mylist1
  } else {
    mylist <- mylist2
  }
  outdt <- sapply(mylist, function(x) sapply(x, mean))
  rownames(outdt) <- methods
  colnames(outdt) <- sample_sizes
  plot(1:dim(outdt)[2], ylim = c(0, ifelse(i == 1, 0.4, 1)), col = "white", ylab = ifelse(i == 1, "Power", "FPR"), 
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty= 'n', cex.lab = 1.3, mgp=c(4, 1, 0))
  title(paste("para=", ifelse(i == 1, "(0.8,0.2)", "(0.5,0.5), FDR=0.1")))
  yticks_val <- seq(0, ifelse(i == 1, 0.4, 1), 0.1)
  axis(2, at = yticks_val, lab = paste0(yticks_val*100, "%"), las = 2, cex.axis = 1.3)
  axis(1, at = 1:dim(outdt)[2], lab = colnames(outdt), cex.axis = 1.3)
  grid()
  for (r in 1:nrow(outdt)) {
    y <- outdt[r, ]
    points(1:dim(outdt)[2], y, col = mycol[r], pch = 19, cex = 1.8)
    lines(1:dim(outdt)[2], y, col = mycol[r], lwd=1.5)
  }
  if (i == 1) {
    legend("topleft", legend = methods, 
           col = mycol[1:length(methods)], lwd = 1, bty = "n", lty=rep(NA,5), pch=rep(19,5))
    # legend("topleft", legend = methods, 
    #        col = mycol[1:length(methods)], lwd = 1, bty = "n", lty=rep(NA,5), pch=rep(19,5), cex = 1.3)
  }
}
# dev.off()
```



### para=c(0.9,0.1)

```{r}
Nsim <- 100
rr <- 0.1
mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61","#FFFF00", "#2b2930")
methods <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
sample_sizes <- c(200, 400, 600, 800, 1000, 1200, 1500)
mylist1 <- mylist2 <- replicate(length(sample_sizes), replicate(length(methods), c(), simplify = FALSE), simplify = FALSE)
m.pvalue <- lapply(1:length(methods), function(x) matrix(nrow = 200, ncol = Nsim))

for (i1 in 0) {
  n <- 0
  for (i3 in sample_sizes) {
    n <- n + 1
    for (i2 in 1:200) {
     load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/TSG0.9/TSG_gene/diff/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
     load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/TSG0.9/TSG_gene/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
     load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/TSG0.9/TSG_gene/other/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
     load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/TSG0.9/TSG_gene/diffFixFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))

      m.pvalue[[1]][i2, ] <- simuresdiffFix[["m1.pvalue"]]
      m.pvalue[[2]][i2, ] <- simuresother[["m1.pvalue"]]
      m.pvalue[[3]][i2, ] <- simuresother[["m2.pvalue"]]
      m.pvalue[[4]][i2, ] <- simuresother[["m3.pvalue"]]
      m.pvalue[[5]][i2, ] <- simuresother[["m4.pvalue"]]
    }
    for (j2 in 1:Nsim) {
      for (k in 1:length(methods)) {
        index1 <- which(p.adjust(m.pvalue[[k]][, j2], method = "BH") < rr)
        sensitivity <- ifelse(length(index1) != 0, length(which(index1 < 51)) / 50, 0)
        specificity <- ifelse(length(index1) != 0, length(which(index1 > 50)) / length(index1), 0)
        mylist1[[n]][[k]] <- c(sensitivity, mylist1[[n]][[k]])
        mylist2[[n]][[k]] <- c(specificity, mylist2[[n]][[k]])
      }
    }
  }
}

# pdf("/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/paper/2.b.pdf", width = 7.8, height = 5.5)
par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))
for (i in 1:2) {
  if (i == 1){
    mylist <- mylist1
  } else {
    mylist <- mylist2
  }
  outdt <- sapply(mylist, function(x) sapply(x, mean))
  rownames(outdt) <- methods
  colnames(outdt) <- sample_sizes
  plot(1:dim(outdt)[2], ylim = c(0, ifelse(i == 1, 0.4, 1)), col = "white", ylab = ifelse(i == 1, "Power", "FPR"), 
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty= 'n', cex.lab = 1.3, mgp=c(4, 1, 0))
  title(paste("para=", ifelse(i == 1, "(0.9,0.1)", "(0.5,0.5), FDR=0.1")))
  yticks_val <- seq(0, ifelse(i == 1, 0.4, 1), 0.1)
  axis(2, at = yticks_val, lab = paste0(yticks_val*100, "%"), las = 2, cex.axis=1.3)
  axis(1, at = 1:dim(outdt)[2], lab = colnames(outdt), cex.axis=1.3)
  grid()
  for (r in 1:nrow(outdt)) {
    y <- outdt[r, ]
    points(1:dim(outdt)[2], y, col = mycol[r], pch = 19, cex = 1.8)
    lines(1:dim(outdt)[2], y, col = mycol[r], lwd=1.5)
  }
  # legend("topleft", legend = methods, 
  #        col = mycol[1:length(methods)], lwd = 1, bty = "n", lty = NA, pch = 19)
}
# dev.off()
```



## Results 3: TSG and continuous phenotype, diffdriver has high power and low FDR


For 200 genes, using five functional annotations, we compare the sensitivity and false discovery rate of different algorithms using binary phenotype based on robust brglm algorithm.  The first 50 genes have differential selection while the left 150 genes do not. The parameter are

`beta_f0=0,1`, `beta_gc=(0,2,0,1)`.

For functional effects,old TSG is  drivermap's TSG parameterfor   

`(functypecode8,       mycons,         sift,    phylop100,           MA, intercept)= (0.45817859, 0.13676461, 0.12436937, 0.08272142, 0.30626015, 0.85127726)`  , while new BRCA's TSG parameter is `((Intercept), functypecode8,        mycons,          sift,     phylop100)=(0.8512773,     0.6136798,     0.3753932,     0.1661389,     0.358036,    0.2229025)` .  The diffdrdiver model use new TSG parameter to simulate and infer while diffdriverFix use new parameter to simulate data and old parameter to select genes, diffdriverFix2 use old parameter to simulate data and infer parameter. 




```{r}
Nsim=50
rr=0.1
mylist1 <- list()
mylist2 <- list()
m1.pvalue=matrix(nrow = 200,ncol = Nsim)
m2.pvalue=matrix(nrow = 200,ncol = Nsim)
m3.pvalue=matrix(nrow = 200,ncol = Nsim)
m4.pvalue=matrix(nrow = 200,ncol = Nsim)
m5.pvalue=matrix(nrow = 200,ncol = Nsim)
m6.pvalue=matrix(nrow = 200,ncol = Nsim)
m7.pvalue=matrix(nrow=200,ncol=Nsim)
m1.fdr=matrix(nrow = 200,ncol = Nsim)
m2.fdr=matrix(nrow = 200,ncol = Nsim)
m3.fdr=matrix(nrow = 200,ncol = Nsim)
m4.fdr=matrix(nrow = 200,ncol = Nsim)
m5.fdr=matrix(nrow = 200,ncol = Nsim)
m6.fdr=matrix(nrow = 200,ncol = Nsim)
m7.fdr=matrix(nrow = 200,ncol = Nsim)

for (i1 in c(0)){
    n<-0
    for(i3 in c(200,400,600, 800,1000,1200)){
      n <- n+1
	for (i2 in 1:200){
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2/diff/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2/other/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))


       ##
     sensitivity1=c()
     specificity1=c()
     sensitivity2=c()
     specificity2=c()
     sensitivity3=c()
     specificity3=c()
     sensitivity4=c()
     specificity4=c()
     sensitivity5=c()
     specificity5=c()
     sensitivity6=c()
     specificity6=c()
     sensitivity7=c()
     specificity7=c()
           
              m1.pvalue[i2,]=simuresother[["m1.pvalue"]]
              m2.pvalue[i2,]=simuresother[["m2.pvalue"]]
              m3.pvalue[i2,]=simuresother[["m3.pvalue"]]
              m4.pvalue[i2,]=simuresother[["m4.pvalue"]]
              m5.pvalue[i2,]=simuresdiffFix[["m1.pvalue"]]
              m6.pvalue[i2,]=simuresdiff[["m1.pvalue"]]

            
}
     for (j2 in 1:Nsim) {
         index1=which(p.adjust(m1.pvalue[,j2],method = "BH")<rr)
            sensitivity1=c(sensitivity1,length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity1=c(specificity1,length(which(index1>50))/length(index1))
            }else{
     specificity1=c(specificity1,0)
}
            
             index1=which(p.adjust(m2.pvalue[,j2],method = "BH")<rr)
             sensitivity2=c(sensitivity2,length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity2=c(specificity2,length(which(index1>50))/length(index1))
            }else{
     specificity2=c(specificity2,0)
}           

           
             index1=which(p.adjust(m3.pvalue[,j2],method = "BH")<rr)
            sensitivity3=c(sensitivity3, length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity3=c(specificity3,length(which(index1>50))/length(index1))
            }else{
     specificity3=c(specificity3,0)
}



             index1=which(p.adjust(m4.pvalue[,j2],method = "BH")<rr)
            sensitivity4=c(sensitivity4,length(which(index1<51))/50)
             if (length(index1)!=0){
                     specificity4=c(specificity4,length(which(index1>50))/length(index1))
            }else{
     specificity4=c(specificity4,0)
}

            
               index1=which(p.adjust(m5.pvalue[,j2],method = "BH")<rr)
             sensitivity5=c(sensitivity5, length(which(index1<51))/50)
               if (length(index1)!=0){
                     specificity5=c(specificity5,length(which(index1>50))/length(index1))
            }else{
     specificity5=c(specificity5,0)
}
            
             index1=which(p.adjust(m6.pvalue[,j2],method = "BH")<rr)
                 sensitivity6=c(sensitivity6,length(which(index1<51))/50)
      if (length(index1)!=0){
                     specificity6=c(specificity6,length(which(index1>50))/length(index1))
            }else{
     specificity6=c(specificity6,0)
}

          }

     
    mylist1[[n]] <- c(mean(sensitivity5), mean(sensitivity1), mean(sensitivity2), mean(sensitivity3), mean(sensitivity4))
    mylist2[[n]] <- c(mean(specificity5), mean(specificity1), mean(specificity2), mean(specificity3), mean(specificity4))
    }

    outdt <- do.call(cbind, mylist1)
#pdf("brglm5.pdf")   
    rownames(outdt)=c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
    colnames(outdt) <- c(200,400, 600,800,1000,1200)
    mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61")
 # mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#fdae61","#FF3D2E")
x=1:dim(outdt)[2] 

# pdf("/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/paper/3.a.pdf", width = 7.5, height = 5.5)
par(mfrow=c(1,2), mar=c(5, 6, 4, 2))

     plot(x,ylim=c(0,0.6), col="white", ylab = "Power", xlab="Number of samples", yaxt="n", xaxt="n", bty="n", cex.lab=1.3, mgp=c(4, 1, 0))
      # title(paste("f0=",i1,",", "beta_gc=(",i2,",","1.513", ")", "P(E=1) =0.5", "P(positive selection|E=1)=0.8", "P(positive selection|E=1)=0.2", sep=" "))
    title( paste( "beta_0=",i1,",","para=(0,2,0,2)",  sep=" "))
    #yticks_val <- pretty(outdt)
    yticks_val <- seq(0,0.6,0.1)
    #axis(2, lab=paste0(yticks_val/Nsim * 100, " %"),las=2)
    axis(2, at=yticks_val, lab=paste0(yticks_val*100, "%"),las=2, cex.axis=1.3)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt), cex.axis=1.3)
    grid()
    for (r in 1:5){
y=outdt[r,]
#y.sd=sddt[r,]
      points(x,y,col=mycol[r], pch=19, cex=1.8)
      lines(x,y,col=mycol[r], lwd=1.5)
 #    arrows(x0=x, y0=y-y.sd, x1=x, y1=y+y.sd, code=3, angle=90, length=0.1,col=mycol[r])
    }
    legend("topleft", legend = c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR"), col= mycol[1:5], lwd=1, bty="n", lty=rep(NA,5), pch=rep(19,5))
    
    
    
     outdt <- do.call(cbind, mylist2)
    
     rownames(outdt)=c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
    colnames(outdt) <- c(200,400, 600,800,1000,1200)
    mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61")
 # mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#fdae61","#FF3D2E")
     plot(1:dim(outdt)[2],ylim=c(0,1), col="white", ylab = "FPR", xlab="Number of samples", yaxt="n", xaxt="n", bty="n", cex.lab=1.3, mgp=c(4, 1, 0))
      # title(paste("f0=",i1,",", "beta_gc=(",i2,",","1.513", ")", "P(E=1) =0.5", "P(positive selection|E=1)=0.8", "P(positive selection|E=1)=0.2", sep=" "))
     title( "para=(0,2,0,0), FDR=0.1")
    #yticks_val <- pretty(outdt)
    yticks_val <- seq(0,1,0.1)
    #axis(2, lab=paste0(yticks_val/Nsim * 100, " %"),las=2)
    axis(2, at=yticks_val, lab=paste0(yticks_val*100, "%"),las=2, cex.axis=1.3)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt), cex.axis=1.3)
    grid()
    for (r in 1:5){
y=outdt[r,]
#y.sd=sddt[r,]
      points(x,y,col=mycol[r], pch=19, cex=1.8)
      lines(x,y,col=mycol[r], lwd=1.5)
#	arrows(x0=x, y0=y-y.sd, x1=x, y1=y+y.sd, code=3, angle=90, length=0.1,col=mycol[r])
    }
    
}
# dev.off()
```


```{r}
Nsim=50
rr=0.1
mylist1 <- list()
mylist2 <- list()
m1.pvalue=matrix(nrow = 200,ncol = Nsim)
m2.pvalue=matrix(nrow = 200,ncol = Nsim)
m3.pvalue=matrix(nrow = 200,ncol = Nsim)
m4.pvalue=matrix(nrow = 200,ncol = Nsim)
m5.pvalue=matrix(nrow = 200,ncol = Nsim)
m6.pvalue=matrix(nrow = 200,ncol = Nsim)
m7.pvalue=matrix(nrow=200,ncol=Nsim)
m1.fdr=matrix(nrow = 200,ncol = Nsim)
m2.fdr=matrix(nrow = 200,ncol = Nsim)
m3.fdr=matrix(nrow = 200,ncol = Nsim)
m4.fdr=matrix(nrow = 200,ncol = Nsim)
m5.fdr=matrix(nrow = 200,ncol = Nsim)
m6.fdr=matrix(nrow = 200,ncol = Nsim)
m7.fdr=matrix(nrow = 200,ncol = Nsim)

for (i1 in c(1)){
    n<-0
    for(i3 in c(200,400,600, 800,1000,1200)){
      n <- n+1
	for (i2 in 1:200){
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2/diff/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/conti2/other/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))


       ##
     sensitivity1=c()
     specificity1=c()
     sensitivity2=c()
     specificity2=c()
     sensitivity3=c()
     specificity3=c()
     sensitivity4=c()
     specificity4=c()
     sensitivity5=c()
     specificity5=c()
     sensitivity6=c()
     specificity6=c()
     sensitivity7=c()
     specificity7=c()
           
              m1.pvalue[i2,]=simuresother[["m1.pvalue"]]
              m2.pvalue[i2,]=simuresother[["m2.pvalue"]]
              m3.pvalue[i2,]=simuresother[["m3.pvalue"]]
              m4.pvalue[i2,]=simuresother[["m4.pvalue"]]
              m5.pvalue[i2,]=simuresdiffFix[["m1.pvalue"]]
              m6.pvalue[i2,]=simuresdiff[["m1.pvalue"]]

            
}
     for (j2 in 1:Nsim) {
         index1=which(p.adjust(m1.pvalue[,j2],method = "BH")<rr)
            sensitivity1=c(sensitivity1,length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity1=c(specificity1,length(which(index1>50))/length(index1))
            }else{
     specificity1=c(specificity1,0)
}
            
             index1=which(p.adjust(m2.pvalue[,j2],method = "BH")<rr)
             sensitivity2=c(sensitivity2,length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity2=c(specificity2,length(which(index1>50))/length(index1))
            }else{
     specificity2=c(specificity2,0)
}           

           
             index1=which(p.adjust(m3.pvalue[,j2],method = "BH")<rr)
            sensitivity3=c(sensitivity3, length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity3=c(specificity3,length(which(index1>50))/length(index1))
            }else{
     specificity3=c(specificity3,0)
}



             index1=which(p.adjust(m4.pvalue[,j2],method = "BH")<rr)
            sensitivity4=c(sensitivity4,length(which(index1<51))/50)
             if (length(index1)!=0){
                     specificity4=c(specificity4,length(which(index1>50))/length(index1))
            }else{
     specificity4=c(specificity4,0)
}

            
               index1=which(p.adjust(m5.pvalue[,j2],method = "BH")<rr)
             sensitivity5=c(sensitivity5, length(which(index1<51))/50)
               if (length(index1)!=0){
                     specificity5=c(specificity5,length(which(index1>50))/length(index1))
            }else{
     specificity5=c(specificity5,0)
}
            
             index1=which(p.adjust(m6.pvalue[,j2],method = "BH")<rr)
                 sensitivity6=c(sensitivity6,length(which(index1<51))/50)
      if (length(index1)!=0){
                     specificity6=c(specificity6,length(which(index1>50))/length(index1))
            }else{
     specificity6=c(specificity6,0)
}

          }

     
    mylist1[[n]] <- c(mean(sensitivity5), mean(sensitivity1), mean(sensitivity2), mean(sensitivity3), mean(sensitivity4))
    mylist2[[n]] <- c(mean(specificity5), mean(specificity1), mean(specificity2), mean(specificity3), mean(specificity4))
    }

    outdt <- do.call(cbind, mylist1)
#pdf("brglm5.pdf")   
    rownames(outdt)=c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
    colnames(outdt) <- c(200,400, 600,800,1000,1200)
    mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61")
 # mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#fdae61","#FF3D2E")
x=1:dim(outdt)[2]   

# pdf("/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/paper/3.b.pdf", width = 7.5, height = 5.5)
par(mfrow=c(1,2), mar=c(5, 6, 4, 2))

     plot(x,ylim=c(0,0.6), col="white", ylab = "Power", xlab="Number of samples", yaxt="n", xaxt="n", bty="n", cex.lab=1.3, mgp=c(4, 1, 0))
      # title(paste("f0=",i1,",", "beta_gc=(",i2,",","1.513", ")", "P(E=1) =0.5", "P(positive selection|E=1)=0.8", "P(positive selection|E=1)=0.2", sep=" "))
     title( paste( "beta_0=",i1,",","para=(0,2,0,2)",  sep=" "))
    #yticks_val <- pretty(outdt)
    yticks_val <- seq(0,1,0.1)
    #axis(2, lab=paste0(yticks_val/Nsim * 100, " %"),las=2)
    axis(2, at=yticks_val, lab=paste0(yticks_val*100, "%"),las=2, cex.axis=1.3)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt), cex.axis=1.3)
    grid()
    for (r in 1:5){
y=outdt[r,]
#y.sd=sddt[r,]
      points(x,y,col=mycol[r], pch=19, cex=1.8)
      lines(x,y,col=mycol[r], lwd=1.5)
 #    arrows(x0=x, y0=y-y.sd, x1=x, y1=y+y.sd, code=3, angle=90, length=0.1,col=mycol[r])
    }
    # legend("topleft", legend = c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR"), col= mycol[1:5], lwd=1, bty="n", lty=rep(NA,5), pch=rep(19,5))
    
    
    
     outdt <- do.call(cbind, mylist2)
    
     rownames(outdt)=c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
    colnames(outdt) <- c(200,400, 600,800,1000,1200)
    mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61")
 # mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#fdae61","#FF3D2E")
     plot(1:dim(outdt)[2],ylim=c(0,1), col="white", ylab = "FPR", xlab="Number of samples", yaxt="n", xaxt="n", bty="n", cex.lab=1.3, mgp=c(4, 1, 0))
      # title(paste("f0=",i1,",", "beta_gc=(",i2,",","1.513", ")", "P(E=1) =0.5", "P(positive selection|E=1)=0.8", "P(positive selection|E=1)=0.2", sep=" "))
     title( "para=(0,2,0,0), FDR=0.1")
    #yticks_val <- pretty(outdt)
    yticks_val <- seq(0,1,0.1)
    #axis(2, lab=paste0(yticks_val/Nsim * 100, " %"),las=2)
    axis(2, at=yticks_val, lab=paste0(yticks_val*100, "%"),las=2, cex.axis=1.3)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt), cex.axis=1.3)
    grid()
    for (r in 1:5){
y=outdt[r,]
#y.sd=sddt[r,]
      points(x,y,col=mycol[r], pch=19, cex=1.8)
      lines(x,y,col=mycol[r], lwd=1.5)
#	arrows(x0=x, y0=y-y.sd, x1=x, y1=y+y.sd, code=3, angle=90, length=0.1,col=mycol[r])
    }
    
}

# dev.off()
```



## Results 4: OG and binary phenotype, diffdriver has high power and low FDR


 For 200 genes, using five functional annotatio:ns,    we compare the sensitivity and false discovery rate of different algorithms using binary phenotype based on robust brglm algorithm.  The first 50 genes have differential selection while the left 150 genes do not. The parameter are
`beta_f0=0,1`, `beta_gc=(0.8,0.2)`.For functional effects,old OG is  drivermap's OG parameterfor 
  `(functypecode8,       mycons,         sift,    phylop100,           MA, intercept)= (-0.18617161,    0.31563535,    0.01956273,    0.25640862,    0.13973529, 0.44973800)`  , while new BRCA's OG parameter is `((Intercept), functypecode8,        mycons,          sift,     phylop100)=`(0.44973800,    0.04907636,    0.80603212,   -0.07773702,    0.66444635,  0.08851684)` .  The diffdrdiver model use new OG parameter to simulate and infer while diffdriverFix use new parameter to simulate data and old parameter to select genes, diffdriverFix2 use old parameter to simulate data and infer parameter. 

### para=c(0.8,0.2)

```{r}
Nsim <- 100
rr <- 0.1
mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61","#FFFF00", "#2b2930")
methods <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
sample_sizes <- c(200, 400, 600, 800, 1000, 1200, 1500)
mylist1 <- mylist2 <- replicate(length(sample_sizes), replicate(length(methods), c(), simplify = FALSE), simplify = FALSE)
m.pvalue <- lapply(1:length(methods), function(x) matrix(nrow = 200, ncol = Nsim))

for (i1 in 0.8) {
  n <- 0
  for (i3 in sample_sizes) {
    n <- n + 1
    for (i2 in 1:200) {
      load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/binaryEM/OG_gene/diff/pppower_betaf0=0_sample",i3,"_",i1,"_",i2,".Rd"))
      load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/binaryEM/OG_gene/diffFix/pppower_betaf0=0_sample",i3,"_",i1,"_",i2,".Rd"))
      load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/binaryEM/OG_gene/other/pppower_betaf0=0_sample",i3,"_",i1,"_",i2,".Rd"))
     
      m.pvalue[[1]][i2, ] <- simuresdiffFix[["m1.pvalue"]]
      m.pvalue[[2]][i2, ] <- simuresother[["m1.pvalue"]]
      m.pvalue[[3]][i2, ] <- simuresother[["m2.pvalue"]]
      m.pvalue[[4]][i2, ] <- simuresother[["m3.pvalue"]]
      m.pvalue[[5]][i2, ] <- simuresother[["m4.pvalue"]]
      
    }
    for (j2 in 1:Nsim) {
      for (k in 1:length(methods)) {
        index1 <- which(p.adjust(m.pvalue[[k]][, j2], method = "BH") < rr)
        sensitivity <- ifelse(length(index1) != 0, length(which(index1 < 51)) / 50, 0)
        specificity <- ifelse(length(index1) != 0, length(which(index1 > 50)) / length(index1), 0)
        mylist1[[n]][[k]] <- c(sensitivity, mylist1[[n]][[k]])
        mylist2[[n]][[k]] <- c(specificity, mylist2[[n]][[k]])
      }
    }
  }
}

# pdf("/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/paper/4.a.pdf", width = 7.8, height = 5.5)
par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))
for (i in 1:2) {
  if (i == 1){
    mylist <- mylist1
  } else {
    mylist <- mylist2
  }
  outdt <- sapply(mylist, function(x) sapply(x, mean))
  rownames(outdt) <- methods
  colnames(outdt) <- sample_sizes
  plot(1:dim(outdt)[2], ylim = c(0, ifelse(i == 1, 0.5, 1)), col = "white", ylab = ifelse(i == 1, "Power", "FPR"), 
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty= 'n', cex.lab = 1.3, mgp=c(4, 1, 0))
  title(paste("para=", ifelse(i == 1, "(0.8,0.2)", "(0.5,0.5), FDR=0.1")))
  yticks_val <- seq(0, ifelse(i == 1, 0.5, 1), 0.1)
  axis(2, at = yticks_val, lab = paste0(yticks_val*100, "%"), las = 2, cex.axis=1.3)
  axis(1, at = 1:dim(outdt)[2], lab = colnames(outdt), cex.axis=1.3)
  grid()
  for (r in 1:nrow(outdt)) {
    y <- outdt[r, ]
    points(1:dim(outdt)[2], y, col = mycol[r], pch = 19, cex = 1.8, cex.axis=1.3)
    lines(1:dim(outdt)[2], y, col = mycol[r], lwd=1.5, cex.axis=1.5)
  }
  if (i == 1) {
  legend("topleft", legend = methods, 
         col = mycol[1:length(methods)], lwd = 1, bty = "n", lty = NA, pch = 19)
  }
}
# dev.off()
```


### para=c(0.9,0.1)

```{r}
Nsim <- 100
rr <- 0.1
mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61","#FFFF00", "#2b2930")
methods <- c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
sample_sizes <- c(200, 400, 600, 800, 1000, 1200, 1500)
mylist1 <- mylist2 <- replicate(length(sample_sizes), replicate(length(methods), c(), simplify = FALSE), simplify = FALSE)
m.pvalue <- lapply(1:length(methods), function(x) matrix(nrow = 200, ncol = Nsim))

for (i1 in 0.9) {
  n <- 0
  for (i3 in sample_sizes) {
    n <- n + 1
    for (i2 in 1:200) {
      load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/binaryEM/OG_gene/diff/pppower_betaf0=0_sample",i3,"_",i1,"_",i2,".Rd"))
      load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/binaryEM/OG_gene/diffFix/pppower_betaf0=0_sample",i3,"_",i1,"_",i2,".Rd"))
      load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/binaryEM/OG_gene/other/pppower_betaf0=0_sample",i3,"_",i1,"_",i2,".Rd"))


      m.pvalue[[1]][i2, ] <- simuresdiffFix[["m1.pvalue"]]
      m.pvalue[[2]][i2, ] <- simuresother[["m1.pvalue"]]
      m.pvalue[[3]][i2, ] <- simuresother[["m2.pvalue"]]
      m.pvalue[[4]][i2, ] <- simuresother[["m3.pvalue"]]
      m.pvalue[[5]][i2, ] <- simuresother[["m4.pvalue"]]
      
    }
    for (j2 in 1:Nsim) {
      for (k in 1:length(methods)) {
        index1 <- which(p.adjust(m.pvalue[[k]][, j2], method = "BH") < rr)
        sensitivity <- ifelse(length(index1) != 0, length(which(index1 < 51)) / 50, 0)
        specificity <- ifelse(length(index1) != 0, length(which(index1 > 50)) / length(index1), 0)
        mylist1[[n]][[k]] <- c(sensitivity, mylist1[[n]][[k]])
        mylist2[[n]][[k]] <- c(specificity, mylist2[[n]][[k]])
      }
    }
  }
}

# pdf("/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/paper/4.b.pdf", width = 7.8, height = 5.5)
par(mfrow = c(1, 2), mar = c(5, 6, 4, 2))
for (i in 1:2) {
  if (i == 1){
    mylist <- mylist1
  } else {
    mylist <- mylist2
  }
  outdt <- sapply(mylist, function(x) sapply(x, mean))
  rownames(outdt) <- methods
  colnames(outdt) <- sample_sizes
  plot(1:dim(outdt)[2], ylim = c(0, ifelse(i == 1, 0.5, 1)), col = "white", ylab = ifelse(i == 1, "Power", "FPR"), 
       xlab = "Number of samples", yaxt = "n", xaxt = "n", bty= 'n', cex.lab = 1.3, mgp=c(4, 1, 0))
  title(paste("para=", ifelse(i == 1, "(0.9,0.1)", "(0.5,0.5), FDR=0.1")))
  yticks_val <- seq(0, ifelse(i == 1, 0.5, 1), 0.1)
  axis(2, at = yticks_val, lab = paste0(yticks_val*100, "%"), las = 2, cex.axis=1.3)
  axis(1, at = 1:dim(outdt)[2], lab = colnames(outdt), cex.axis=1.3)
  grid()
  for (r in 1:nrow(outdt)) {
    y <- outdt[r, ]
    points(1:dim(outdt)[2], y, col = mycol[r], pch = 19, cex = 1.8)
    lines(1:dim(outdt)[2], y, col = mycol[r], lwd=1.5)
  }
  # legend("topleft", legend = methods, 
  #        col = mycol[1:length(methods)], lwd = 1, bty = "n", lty = NA, pch = 19)
}
# dev.off()
```


## Results 5: OG and continuous phenotype, diffdriver has high power and low FDR

For 200 genes, using five functional annotatio:ns,    we compare the sensitivity and false discovery rate of different algorithms using binary phenotype based on robust brglm algorithm.  The first 50 genes have differential selection while the left 150 genes do not. The parameter are

`beta_f0=0,1`, `beta_gc=(0,2,0,1)`. 
FFor functional effects,old OG is  drivermap's OG parameterfor   `(functypecode8,       mycons,         sift,    phylop100,           MA, intercept)= (0.44973800,    0.04907636, 0.80603212, -0.07773702, 0.66444635, 0.08851684)`  , while new BRCA's OG parameter is `((Intercept), functypecode8,        mycons,          sift,     phylop100)=`(-0.18617161,0.31563535,0.01956273,0.25640862,0.13973529,0.44973800)` . 
The diffdrdiver model use new OG parameter to simulate and infer while diffdriverFix use new parameter to simulate data and old parameter to select genes, diffdriverFix2 use old parameter to simulate data and infer parameter. 


```{r}
Nsim=50
rr=0.1
mylist1 <- list()
mylist2 <- list()
m1.pvalue=matrix(nrow = 200,ncol = Nsim)
m2.pvalue=matrix(nrow = 200,ncol = Nsim)
m3.pvalue=matrix(nrow = 200,ncol = Nsim)
m4.pvalue=matrix(nrow = 200,ncol = Nsim)
m5.pvalue=matrix(nrow = 200,ncol = Nsim)
m6.pvalue=matrix(nrow = 200,ncol = Nsim)
m7.pvalue=matrix(nrow=200,ncol=Nsim)
m1.fdr=matrix(nrow = 200,ncol = Nsim)
m2.fdr=matrix(nrow = 200,ncol = Nsim)
m3.fdr=matrix(nrow = 200,ncol = Nsim)
m4.fdr=matrix(nrow = 200,ncol = Nsim)
m5.fdr=matrix(nrow = 200,ncol = Nsim)
m6.fdr=matrix(nrow = 200,ncol = Nsim)
m7.fdr=matrix(nrow = 200,ncol = Nsim)



for (i1 in c(0)){
    n<-0
    for(i3 in c(200,400,600, 800,1000,1200)){
      n <- n+1
	for (i2 in 1:200){
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene/diff/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene/other/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))


       ##
     sensitivity1=c()
     specificity1=c()
     sensitivity2=c()
     specificity2=c()
     sensitivity3=c()
     specificity3=c()
     sensitivity4=c()
     specificity4=c()
     sensitivity5=c()
     specificity5=c()
     sensitivity6=c()
     specificity6=c()
     sensitivity7=c()
     specificity7=c()
           
              m1.pvalue[i2,]=simuresother[["m1.pvalue"]]
              m2.pvalue[i2,]=simuresother[["m2.pvalue"]]
              m3.pvalue[i2,]=simuresother[["m3.pvalue"]]
              m4.pvalue[i2,]=simuresother[["m4.pvalue"]]
              m5.pvalue[i2,]=simuresdiffFix[["m1.pvalue"]]
              m6.pvalue[i2,]=simuresdiff[["m1.pvalue"]]

            
}
     for (j2 in 1:Nsim) {
         index1=which(p.adjust(m1.pvalue[,j2],method = "BH")<rr)
            sensitivity1=c(sensitivity1,length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity1=c(specificity1,length(which(index1>50))/length(index1))
            }else{
     specificity1=c(specificity1,0)
}
            
             index1=which(p.adjust(m2.pvalue[,j2],method = "BH")<rr)
             sensitivity2=c(sensitivity2,length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity2=c(specificity2,length(which(index1>50))/length(index1))
            }else{
     specificity2=c(specificity2,0)
}           

           
             index1=which(p.adjust(m3.pvalue[,j2],method = "BH")<rr)
            sensitivity3=c(sensitivity3, length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity3=c(specificity3,length(which(index1>50))/length(index1))
            }else{
     specificity3=c(specificity3,0)
}



             index1=which(p.adjust(m4.pvalue[,j2],method = "BH")<rr)
            sensitivity4=c(sensitivity4,length(which(index1<51))/50)
             if (length(index1)!=0){
                     specificity4=c(specificity4,length(which(index1>50))/length(index1))
            }else{
     specificity4=c(specificity4,0)
}

            
               index1=which(p.adjust(m5.pvalue[,j2],method = "BH")<rr)
             sensitivity5=c(sensitivity5, length(which(index1<51))/50)
               if (length(index1)!=0){
                     specificity5=c(specificity5,length(which(index1>50))/length(index1))
            }else{
     specificity5=c(specificity5,0)
}
            
             index1=which(p.adjust(m6.pvalue[,j2],method = "BH")<rr)
                 sensitivity6=c(sensitivity6,length(which(index1<51))/50)
      if (length(index1)!=0){
                     specificity6=c(specificity6,length(which(index1>50))/length(index1))
            }else{
     specificity6=c(specificity6,0)
}

          }


    mylist1[[n]] <- c(mean(sensitivity5), mean(sensitivity1), mean(sensitivity2), mean(sensitivity3), mean(sensitivity4))
    mylist2[[n]] <- c(mean(specificity5), mean(specificity1), mean(specificity2), mean(specificity3), mean(specificity4))
    
    }

    outdt <- do.call(cbind, mylist1)
#pdf("brglm5.pdf")   
    rownames(outdt)=c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
    colnames(outdt) <- c(200,400, 600,800,1000,1200)
    mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61")
 # mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#fdae61","#FF3D2E")
x=1:dim(outdt)[2]  

## Save pdf
# pdf("/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/paper/5.a.pdf", width = 7.5, height = 5.5)

par(mfrow=c(1,2), mar=c(5, 6, 4, 2))
     plot(x,ylim=c(0,0.7), col="white", ylab = "Power", xlab="Number of samples", yaxt="n", xaxt="n", bty="n", cex.lab=1.3, mgp=c(4, 1, 0))
      # title(paste("f0=",i1,",", "beta_gc=(",i2,",","1.513", ")", "P(E=1) =0.5", "P(positive selection|E=1)=0.8", "P(positive selection|E=1)=0.2", sep=" "))
     title( paste( "beta_0=",i1,",","para=(0,2,0,2)",  sep=" "))
    #yticks_val <- pretty(outdt)
    yticks_val <- seq(0,0.7,0.1)
    #axis(2, lab=paste0(yticks_val/Nsim * 100, " %"),las=2)
    axis(2, at=yticks_val, lab=paste0(yticks_val*100, "%"),las=2, cex.axis=1.3)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt), cex.axis=1.3)
    grid()
    for (r in 1:5){
y=outdt[r,]
#y.sd=sddt[r,]
      points(x,y,col=mycol[r], pch=19, cex=1.8)
      lines(x,y,col=mycol[r], lwd=1.5)
 #    arrows(x0=x, y0=y-y.sd, x1=x, y1=y+y.sd, code=3, angle=90, length=0.1,col=mycol[r])
    }
    legend("topleft", legend = c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR"), col= mycol[1:5], lwd=1, bty="n", lty=rep(NA,5), pch=rep(19,5))
    
    
    
     outdt <- do.call(cbind, mylist2)
    
     rownames(outdt)=c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
    colnames(outdt) <- c(200,400, 600,800,1000,1200)
    mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61")
 # mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#fdae61","#FF3D2E")
     plot(1:dim(outdt)[2],ylim=c(0,1), col="white", ylab = "FPR", xlab="Number of samples", yaxt="n", xaxt="n", bty="n", cex.lab=1.3, mgp=c(4, 1, 0))
      # title(paste("f0=",i1,",", "beta_gc=(",i2,",","1.513", ")", "P(E=1) =0.5", "P(positive selection|E=1)=0.8", "P(positive selection|E=1)=0.2", sep=" "))
     title( "para=(0,2,0,0), FDR=0.1")
    #yticks_val <- pretty(outdt)
    yticks_val <- seq(0,1,0.1)
    #axis(2, lab=paste0(yticks_val/Nsim * 100, " %"),las=2)
    axis(2, at=yticks_val, lab=paste0(yticks_val*100, "%"),las=2, cex.axis=1.3)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt), cex.axis=1.3)
    grid()
    for (r in 1:5){
y=outdt[r,]
#y.sd=sddt[r,]
      points(x,y,col=mycol[r], pch=19, cex=1.8)
      lines(x,y,col=mycol[r], lwd=1.5)
#	arrows(x0=x, y0=y-y.sd, x1=x, y1=y+y.sd, code=3, angle=90, length=0.1,col=mycol[r])
    }
    # legend("topleft", legend = c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR"), col= mycol[1:5], lwd=1, bty="n", lty=rep(NA,5), pch=rep(19,5))
}

# dev.off()
```



```{r}
Nsim=50
rr=0.1
mylist1 <- list()
mylist2 <- list()
m1.pvalue=matrix(nrow = 200,ncol = Nsim)
m2.pvalue=matrix(nrow = 200,ncol = Nsim)
m3.pvalue=matrix(nrow = 200,ncol = Nsim)
m4.pvalue=matrix(nrow = 200,ncol = Nsim)
m5.pvalue=matrix(nrow = 200,ncol = Nsim)
m6.pvalue=matrix(nrow = 200,ncol = Nsim)
m7.pvalue=matrix(nrow=200,ncol=Nsim)
m1.fdr=matrix(nrow = 200,ncol = Nsim)
m2.fdr=matrix(nrow = 200,ncol = Nsim)
m3.fdr=matrix(nrow = 200,ncol = Nsim)
m4.fdr=matrix(nrow = 200,ncol = Nsim)
m5.fdr=matrix(nrow = 200,ncol = Nsim)
m6.fdr=matrix(nrow = 200,ncol = Nsim)
m7.fdr=matrix(nrow = 200,ncol = Nsim)

for (i1 in c(1)){
    n<-0
    for(i3 in c(200,400,600, 800,1000,1200)){
      n <- n+1
	for (i2 in 1:200){
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene/diff/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/OCG/OG_gene/other/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))


       ##
     sensitivity1=c()
     specificity1=c()
     sensitivity2=c()
     specificity2=c()
     sensitivity3=c()
     specificity3=c()
     sensitivity4=c()
     specificity4=c()
     sensitivity5=c()
     specificity5=c()
     sensitivity6=c()
     specificity6=c()
     sensitivity7=c()
     specificity7=c()
           
              m1.pvalue[i2,]=simuresother[["m1.pvalue"]]
              m2.pvalue[i2,]=simuresother[["m2.pvalue"]]
              m3.pvalue[i2,]=simuresother[["m3.pvalue"]]
              m4.pvalue[i2,]=simuresother[["m4.pvalue"]]
              m5.pvalue[i2,]=simuresdiffFix[["m1.pvalue"]]
              m6.pvalue[i2,]=simuresdiff[["m1.pvalue"]]

            
}
     for (j2 in 1:Nsim) {
         index1=which(p.adjust(m1.pvalue[,j2],method = "BH")<rr)
            sensitivity1=c(sensitivity1,length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity1=c(specificity1,length(which(index1>50))/length(index1))
            }else{
     specificity1=c(specificity1,0)
}
            
             index1=which(p.adjust(m2.pvalue[,j2],method = "BH")<rr)
             sensitivity2=c(sensitivity2,length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity2=c(specificity2,length(which(index1>50))/length(index1))
            }else{
     specificity2=c(specificity2,0)
}           

           
             index1=which(p.adjust(m3.pvalue[,j2],method = "BH")<rr)
            sensitivity3=c(sensitivity3, length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity3=c(specificity3,length(which(index1>50))/length(index1))
            }else{
     specificity3=c(specificity3,0)
}



             index1=which(p.adjust(m4.pvalue[,j2],method = "BH")<rr)
            sensitivity4=c(sensitivity4,length(which(index1<51))/50)
             if (length(index1)!=0){
                     specificity4=c(specificity4,length(which(index1>50))/length(index1))
            }else{
     specificity4=c(specificity4,0)
}

            
               index1=which(p.adjust(m5.pvalue[,j2],method = "BH")<rr)
             sensitivity5=c(sensitivity5, length(which(index1<51))/50)
               if (length(index1)!=0){
                     specificity5=c(specificity5,length(which(index1>50))/length(index1))
            }else{
     specificity5=c(specificity5,0)
}
            
             index1=which(p.adjust(m6.pvalue[,j2],method = "BH")<rr)
                 sensitivity6=c(sensitivity6,length(which(index1<51))/50)
      if (length(index1)!=0){
                     specificity6=c(specificity6,length(which(index1>50))/length(index1))
            }else{
     specificity6=c(specificity6,0)
}

          }

     
    mylist1[[n]] <- c(mean(sensitivity5), mean(sensitivity1), mean(sensitivity2), mean(sensitivity3), mean(sensitivity4))
    mylist2[[n]] <- c(mean(specificity5), mean(specificity1), mean(specificity2), mean(specificity3), mean(specificity4))
    }

    outdt <- do.call(cbind, mylist1)
#pdf("brglm5.pdf")   
    rownames(outdt)=c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
    colnames(outdt) <- c(200,400, 600,800,1000,1200)
    mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61")
 # mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#fdae61","#FF3D2E")
x=1:dim(outdt)[2]   

## Save pdf
# pdf("/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/paper/5.b.pdf", width = 7.5, height = 5.5)

# par(mfrow=c(1,2))
par(mfrow=c(1,2), mar=c(5, 6, 4, 2))

     plot(x,ylim=c(0,0.7), col="white", ylab = "Power", xlab="Number of samples", yaxt="n", xaxt="n", bty="n", cex.lab=1.3, mgp=c(4, 1, 0))
      # title(paste("f0=",i1,",", "beta_gc=(",i2,",","1.513", ")", "P(E=1) =0.5", "P(positive selection|E=1)=0.8", "P(positive selection|E=1)=0.2", sep=" "))
     title( paste( "beta_0=",i1,",","para=(0,2,0,2)",  sep=" "))
    #yticks_val <- pretty(outdt)
    yticks_val <- seq(0,0.7,0.1)
    #axis(2, lab=paste0(yticks_val/Nsim * 100, " %"),las=2)
    axis(2, at=yticks_val, lab=paste0(yticks_val*100, "%"),las=2, cex.axis=1.3)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt), cex.axis=1.3)
    grid()
    for (r in 1:5){
y=outdt[r,]
#y.sd=sddt[r,]
      points(x,y,col=mycol[r], pch=19, cex=1.8)
      lines(x,y,col=mycol[r], lwd=1.5)
 #    arrows(x0=x, y0=y-y.sd, x1=x, y1=y+y.sd, code=3, angle=90, length=0.1,col=mycol[r])
    }
    # legend("topleft", legend = c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR"), col= mycol[1:5], lwd=1, bty="n", lty=rep(NA,5), pch=rep(19,5), cex=1.0)
    
    
    
     outdt <- do.call(cbind, mylist2)
    
     rownames(outdt)=c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
    colnames(outdt) <- c(200,400, 600,800,1000,1200)
    mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61")
 # mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#fdae61","#FF3D2E")
     plot(1:dim(outdt)[2],ylim=c(0,1), col="white", ylab = "FPR", xlab="Number of samples", yaxt="n", xaxt="n", bty="n", cex.lab=1.3, mgp=c(4, 1, 0))
      # title(paste("f0=",i1,",", "beta_gc=(",i2,",","1.513", ")", "P(E=1) =0.5", "P(positive selection|E=1)=0.8", "P(positive selection|E=1)=0.2", sep=" "))
     title( "para=(0,2,0,0), FDR=0.1")
    #yticks_val <- pretty(outdt)
    yticks_val <- seq(0,1,0.1)
    #axis(2, lab=paste0(yticks_val/Nsim * 100, " %"),las=2)
    axis(2, at=yticks_val, lab=paste0(yticks_val*100, "%"),las=2, cex.axis=1.3)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt), cex.axis=1.3)
    grid()
    for (r in 1:5){
y=outdt[r,]
#y.sd=sddt[r,]
      points(x,y,col=mycol[r], pch=19, cex=1.8)
      lines(x,y,col=mycol[r], lwd=1.5)
#	arrows(x0=x, y0=y-y.sd, x1=x, y1=y+y.sd, code=3, angle=90, length=0.1,col=mycol[r])
    }
    # legend("topleft", legend = c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR"), col= mycol[1:5], lwd=1, bty="n", lty=rep(NA,5), pch=rep(19,5))
}

# dev.off()

```



## Results 6: Sensitivity vs Specificity: continuous phenotype with BRCA TSG functional effects

Aim: For 200 genes, using five functional annotatio:ns,    we compare the sensitivity and false discovery rate of different algorithms using binary phenotype based on robust brglm algorithm.  The first 50 genes have differential selection while the left 150 genes do not. The parameter are

`beta_f0=0,1`, `beta_gc=(0,2,0,1)`.

For functional effects,old TSG is  drivermap's TSG parameterfor   

`(functypecode8,       mycons,         sift,    phylop100,           MA, intercept)= (0.45817859, 0.13676461, 0.12436937, 0.08272142, 0.30626015, 0.85127726)`  , while new BRCA's TSG parameter is `((Intercept), functypecode8,        mycons,          sift,     phylop100)=(0.8512773,     0.6136798,     0.3753932,     0.1661389,     0.358036,    0.2229025)` .  The diffdrdiver model use new TSG parameter to simulate and infer while diffdriverFix use new parameter to simulate data and old parameter to select genes, diffdriverFix2 use old parameter to simulate data and infer parameter. 

```{r}
Nsim=50
rr=0.1
mylist1 <- list()
mylist2 <- list()
m1.pvalue=matrix(nrow = 200,ncol = Nsim)
m2.pvalue=matrix(nrow = 200,ncol = Nsim)
m3.pvalue=matrix(nrow = 200,ncol = Nsim)
m4.pvalue=matrix(nrow = 200,ncol = Nsim)
m5.pvalue=matrix(nrow = 200,ncol = Nsim)
m6.pvalue=matrix(nrow = 200,ncol = Nsim)
m7.pvalue=matrix(nrow=200,ncol=Nsim)
m1.fdr=matrix(nrow = 200,ncol = Nsim)
m2.fdr=matrix(nrow = 200,ncol = Nsim)
m3.fdr=matrix(nrow = 200,ncol = Nsim)
m4.fdr=matrix(nrow = 200,ncol = Nsim)
m5.fdr=matrix(nrow = 200,ncol = Nsim)
m6.fdr=matrix(nrow = 200,ncol = Nsim)
m7.fdr=matrix(nrow = 200,ncol = Nsim)

for (i1 in c(0,1)){

    mylist1 <- list()
    mylist2 <- list()
    n<-0
    for(i3 in c(200,400,600, 800,1000,1200,1500,2000)){
      n <- n+1
	for (i2 in 1:200){
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/diff/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/diffFix/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))
load(paste0("/dartfs/rc/lab/S/Szhao/jiezhou/diffdata/gene239/conti/TSG_BRCA/conti/other/pppower_betaf0=",i1,"_sample",i3,"_",i2,".Rd"))


       ##
     sensitivity1=c()
     specificity1=c()
     sensitivity2=c()
     specificity2=c()
     sensitivity3=c()
     specificity3=c()
     sensitivity4=c()
     specificity4=c()
     sensitivity5=c()
     specificity5=c()
     sensitivity6=c()
     specificity6=c()
     sensitivity7=c()
     specificity7=c()
           
              m1.pvalue[i2,]=simuresother[["m1.pvalue"]]
              m2.pvalue[i2,]=simuresother[["m2.pvalue"]]
              m3.pvalue[i2,]=simuresother[["m3.pvalue"]]
              m4.pvalue[i2,]=simuresother[["m4.pvalue"]]
              m5.pvalue[i2,]=simuresdiffFix[["m1.pvalue"]]
              m6.pvalue[i2,]=simuresdiff[["m1.pvalue"]]

            
}
     for (j2 in 1:Nsim) {
         index1=which(p.adjust(m1.pvalue[,j2],method = "BH")<rr)
            sensitivity1=c(sensitivity1,length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity1=c(specificity1,length(which(index1>50))/length(index1))
            }else{
     specificity1=c(specificity1,0)
}
            
             index1=which(p.adjust(m2.pvalue[,j2],method = "BH")<rr)
             sensitivity2=c(sensitivity2,length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity2=c(specificity2,length(which(index1>50))/length(index1))
            }else{
     specificity2=c(specificity2,0)
}           

           
             index1=which(p.adjust(m3.pvalue[,j2],method = "BH")<rr)
            sensitivity3=c(sensitivity3, length(which(index1<51))/50)
if (length(index1)!=0){
                     specificity3=c(specificity3,length(which(index1>50))/length(index1))
            }else{
     specificity3=c(specificity3,0)
}



             index1=which(p.adjust(m4.pvalue[,j2],method = "BH")<rr)
            sensitivity4=c(sensitivity4,length(which(index1<51))/50)
             if (length(index1)!=0){
                     specificity4=c(specificity4,length(which(index1>50))/length(index1))
            }else{
     specificity4=c(specificity4,0)
}

            
               index1=which(p.adjust(m5.pvalue[,j2],method = "BH")<rr)
             sensitivity5=c(sensitivity5, length(which(index1<51))/50)
               if (length(index1)!=0){
                     specificity5=c(specificity5,length(which(index1>50))/length(index1))
            }else{
     specificity5=c(specificity5,0)
}
            
             index1=which(p.adjust(m6.pvalue[,j2],method = "BH")<rr)
                 sensitivity6=c(sensitivity6,length(which(index1<51))/50)
      if (length(index1)!=0){
                     specificity6=c(specificity6,length(which(index1>50))/length(index1))
            }else{
     specificity6=c(specificity6,0)
}

          }

     
      mylist1[[n]] <- c(mean(sensitivity5), mean(sensitivity1), mean(sensitivity2), mean(sensitivity3), mean(sensitivity4))
      mylist2[[n]] <- c(mean(specificity5), mean(specificity1), mean(specificity2), mean(specificity3), mean(specificity4))
    }

    outdt <- do.call(cbind, mylist1)
#pdf("brglm5.pdf")   
     rownames(outdt)=c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
    colnames(outdt) <- c(200,400, 600,800,1000,1200,1500,2000)
    mycol=c("#FF3D2E","#00AFBB","#aa00ff","#0001f6","#fdae61")
 # mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#fdae61","#FF3D2E")
x=1:dim(outdt)[2]   

# Save PDF
# pdf(paste0("/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/paper/6.", i1, ".pdf"), width = 7.5, height = 5.5)
par(mfrow=c(1,2), mar=c(5, 6, 4, 2))

     plot(x,ylim=c(0,0.6), col="white", ylab = "Power", xlab="Number of samples", yaxt="n", xaxt="n", bty="n", cex.lab=1.3, mgp=c(4, 1, 0))
      # title(paste("f0=",i1,",", "beta_gc=(",i2,",","1.513", ")", "P(E=1) =0.5", "P(positive selection|E=1)=0.8", "P(positive selection|E=1)=0.2", sep=" "))
     title( paste( "beta_0=",i1,",","para=(0,2,0,1)",  sep=" "))
    #yticks_val <- pretty(outdt)
    yticks_val <- seq(0,0.6,0.1)
    #axis(2, lab=paste0(yticks_val/Nsim * 100, " %"),las=2)
    axis(2, at=yticks_val, lab=paste0(yticks_val*100, "%"),las=2, cex.axis=1.3)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt), cex.axis=1.1)
    grid()
    for (r in 1:5){
y=outdt[r,]
#y.sd=sddt[r,]
      points(x,y,col=mycol[r], pch=19, cex=1.8)
      lines(x,y,col=mycol[r], lwd=1.5)
 #    arrows(x0=x, y0=y-y.sd, x1=x, y1=y+y.sd, code=3, angle=90, length=0.1,col=mycol[r])
    }
    legend("topleft", legend = c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR"), col= mycol[1:5], lwd=1, bty="n", lty=rep(NA,5), pch=rep(19,5))
    
    
    
     outdt <- do.call(cbind, mylist2)
    
     rownames(outdt)=c("DiffDriver", "Linear", "Fisher", "Binomial", "LogisticR")
    colnames(outdt) <- c(200,400, 600,800,1000,1200,1500,2000)
 # mycol=c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#fdae61","#FF3D2E")
     plot(1:dim(outdt)[2],ylim=c(0,1), col="white", ylab = "FPR", xlab="Number of samples", yaxt="n", xaxt="n", bty="n", cex.lab=1.3, mgp=c(4, 1, 0))
      # title(paste("f0=",i1,",", "beta_gc=(",i2,",","1.513", ")", "P(E=1) =0.5", "P(positive selection|E=1)=0.8", "P(positive selection|E=1)=0.2", sep=" "))
     title( "para=(0,2,0,0), FDR=0.1")
    #yticks_val <- pretty(outdt)
    yticks_val <- seq(0,1,0.1)
    #axis(2, lab=paste0(yticks_val/Nsim * 100, " %"),las=2)
    axis(2, at=yticks_val, lab=paste0(yticks_val*100, "%"),las=2, cex.axis=1.3)
    axis(1, at=1:dim(outdt)[2], lab=colnames(outdt), cex.axis=1.1)
    grid()
    for (r in 1:5){
y=outdt[r,]
#y.sd=sddt[r,]
      points(x,y,col=mycol[r], pch=19, cex=1.8)
      lines(x,y,col=mycol[r], lwd=1.5)
#	arrows(x0=x, y0=y-y.sd, x1=x, y1=y+y.sd, code=3, angle=90, length=0.1,col=mycol[r])
    }
    
    # dev.off()
}


```

