---
title: "Simulation: Confounding Between Phenotype and Mutational Signatures"
author: "Jie Zhou, Siming Zhao"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: hide
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This simulation demonstrates that when the phenotype is correlated with
mutational signatures — but **not** with selection — benchmark methods produce
false positives while DiffDriver correctly controls the false positive rate.

### Simulation setup

- **Background mutation rate (realistic, LUAD-derived)**: The per-site
  background probability is `bmr = (signatures %*% loadings) / sc`, built entirely
  from LUAD data — so it is *not* an arbitrary rate:
    - **Shape**: the `signatures` (96 × 6) and per-sample `loadings` (`ll$LUAD`)
      are the real LUAD mutational signatures and exposures (fastTopics on the
      LUAD mutation counts, 527 samples, bootstrapped up to 1,500). The loadings
      are a composition (constant column sums = 62.75), so a summed-loading
      covariate carries no phenotype information and cannot adjust the confounding
      away.
    - **Level**: `sc = 359000` gives a mean per-site probability of ≈1.8×10⁻⁶ and
      a genome-wide **synonymous burden of ≈63 mutations per sample** (median 62,
      range 56–73), matching the **observed ≈64 synonymous mutations per sample**
      in real TCGA-LUAD (530 samples). So both the spectrum and the overall level
      are LUAD-realistic.
- **Mutational signatures**: Six signatures extracted from LUAD using
  [fastTopics](https://stephenslab.github.io/fastTopics/). Signature 3 — a
  T>G–dominated (smoking-like) signature whose highest-loading context is
  `C[T>G]G` (nucleotide type 34) — is selected as the confounding signature.
- **Phenotype**: Simulated to correlate with signature 3 ($R^2 = 0.78$), but
  with **zero correlation** with selection (`para = c(0.5, 0.5)`).
- **Mutations**: Simulated for a single gene (ERBB3) using oncogene mutation
  rates (background + selection), with two hotspots placed at nucleotide type 34
  (the context with the highest loading in signature 3), so that the
  hotspot-dominated mutations track the phenotype.
- **Background estimation (as in real analyses)**: Neither DiffDriver nor
  [coselens](https://github.com/ggonzalezlopez/coselens) is handed the true
  background rate. Both **estimate** it from a simulated genome-wide **silent
  (synonymous) mutation count matrix** calibrated to the LUAD synonymous burden —
  exactly what each tool does on real data (see
  [below](#estimating-the-background-from-silent-mutations)).

### Key result

Because the phenotype correlates with a mutational signature that affects
overall mutation rates, the count-based benchmark methods (Linear regression,
Fisher's exact test, Binomial test, Logistic regression, and the Mann-Whitney
rank-sum test) increasingly detect differential selection as sample size grows —
all false positives. DiffDriver and coselens, which model a signature-based
background mutation rate estimated from the silent spectrum, maintain a low false
positive rate. The count-based benchmarks do **not** use or estimate a background
rate — they operate on the raw mutation counts, which is why they cannot separate
the confounded background from selection.

```{r keyresult, fig.width=7, fig.height=6}
Nsim <- 100
SIZES <- c(200,400,600,800,1000,1200,1500)
load("data/pppower_estimatedBMR_bmr.Rd")     # object `estBMR`, keyed by N (estimated background)
load("data/pppower_mannwhitney_bmr.Rd")      # object `mw`

fp   <- function(p) mean(p < 0.01, na.rm = TRUE)
rate <- function(field) sapply(as.character(SIZES), function(n) fp(estBMR[[n]][[field]]))
cosE <- rate("coselens_est"); ddE <- rate("diff_est")   # estimated-background methods

# count-based benchmarks (do not use a background rate; from the archived runs)
lin <- fis <- bin <- lr <- mw_fp <- numeric(length(SIZES))
for (i in seq_along(SIZES)) {
  N <- SIZES[i]
  o <- new.env(); load(sprintf("data/bmr_archive/other/pppower_betaf0=1_sample%d_0.5_0.Rd", N), envir = o)
  s <- o$simuresother
  lin[i] <- fp(s$m1.pvalue); fis[i] <- fp(s$m2.pvalue); bin[i] <- fp(s$m3.pvalue); lr[i] <- fp(s$m4.pvalue)
  mw_fp[i] <- fp(mw[[as.character(N)]]$mannwhitney.pvalue)
}

outdt <- rbind(Linear = lin, Fisher = fis, Binomial = bin, LogisticR = lr,
               MannWhitney = mw_fp, Coselens = cosE, DiffDriver = ddE)
colnames(outdt) <- SIZES
mycol <- c("#00AFBB","#2b8cbe","#aa00ff","#0001f6","#1a9850","#ff7f00","#FF3D2E")
mypch <- c(19,19,19,19,3,8,17)
x <- seq_along(SIZES)
plot(x, ylim = c(0,1), col = "white", ylab = "False positive rate (p < 0.01)",
     xlab = "Number of samples", yaxt = "n", xaxt = "n")
title("Phenotype–signature confounding (background estimated from silent mutations)")
axis(2, at = seq(0,1,0.1), las = 2); axis(1, at = x, labels = SIZES); grid()
abline(h = 0.01, lty = 3, col = "grey60")
for (r in 1:7) { lines(x, outdt[r,], col = mycol[r]); points(x, outdt[r,], col = mycol[r], pch = mypch[r], cex = 1.3) }
legend("topleft", rownames(outdt), col = mycol, lwd = 1, bty = "n", lty = NA, pch = mypch)
```

DiffDriver and coselens keep the false positive rate near zero even though they
must *estimate* the confounded background from a finite silent spectrum — the
estimation does not reintroduce the confounding.

### Robustness across realistic variations

The [Signature-Confounding under Realistic Variations](Simulation-BMR-variations.html)
page stress-tests this benchmark under variable per-sample mutation burden,
different confounding strengths ($R^2$), a continuous phenotype, and DiffDriver's
robustness to the number of fitted signatures. The count-based benchmarks stay
vulnerable throughout, while DiffDriver and Coselens keep the false positive rate
controlled.

## Rerunning the simulation

The simulation functions live in [`code/simulate_functions.R`](https://github.com/szhaolab/diffdriver-paper/blob/main/code/simulate_functions.R)
under the "CONFOUNDING scenario" section (`build_confounded_bmr`,
`simulate_confounding`, `run_confounding_benchmarks`). The count-based benchmark
methods are all in `run_confounding_benchmarks`
(Linear/Fisher/Binomial/LogisticR + Mann-Whitney).

The inputs are bundled in `data/bmr_sim_inputs.rda` (`beta_gc`, `beta_gcFix`,
`signature`, `ll`, `sganno`, `sgmatrix`), together with `data/hmm.rda`,
`data/codeSignature.rda`, and `data/hotspot1sig.rda` (hotspots at nttype 34).
The count-based benchmarks at a single sample size are reproduced with:

```{r eval=FALSE}
library(diffdriver); library(data.table); library(Matrix)
source("code/simulate_functions.R")
load("data/bmr_sim_inputs.rda"); load("data/hmm.rda")
load("data/codeSignature.rda");  load("data/hotspot1sig.rda")

set.seed(1)
res_other <- run_confounding_benchmarks(
  binary = TRUE, Niter = 100,
  sganno = bmr_sim$sganno, sgmatrix = bmr_sim$sgmatrix,
  Nsample = 1500, para = c(0.5, 0.5),      # no differential selection
  rho = sqrt(0.78),                        # phenotype ~ signature 3, R^2 = 0.78
  signatures = bmr_sim$signature, loadings = bmr_sim$ll,
  beta_gc = bmr_sim$beta_gc, hot = 1, hmm = hmm,
  sc = 359000)                             # background-rate scale (~188 mutations)

# false-positive rate per method (p < 0.01), no true signal so all are FPs
sapply(res_other[c("m1.pvalue","m2.pvalue","m3.pvalue","m4.pvalue",
                    "mannwhitney.pvalue")], function(p) mean(p < 0.01))
```

The full benchmark loops this over `Nsample = 200 … 1500`; the Mann-Whitney
p-values used above were produced this way and saved to
`data/pppower_mannwhitney_bmr.Rd`. Because the original
`scripts/simulate_functions.R` (and the exact HMM coefficients) were lost, the
rerun is *close* to — but not bit-identical to — the archived runs; the Linear
and Fisher/Binomial/LogisticR curves plotted above are read from the original
archived results.

### Estimating the background from silent mutations

DiffDriver and coselens do not receive the true background — each **estimates** it
from a simulated genome-wide **silent (synonymous) mutation count matrix**, exactly
as it would on real data. The silent matrix is not arbitrary — it is simulated from
the same realistic LUAD background:

$$Y[c,s] \sim \text{Poisson}\big(\text{Lsyn96}[c]\cdot \text{bmr}[c,s]\cdot g\big),$$

where `bmr = (signatures %*% loadings) / sc` uses the **real LUAD** signatures
(96 × 6) and per-sample loadings, `Lsyn96` is the number of synonymous sites per
trinucleotide context (dndscv RefCDS; `data/Lsyn.rds`, ≈ 23.3 M sites), and `g`
calibrates the mean per-sample synonymous burden to the **observed ≈ 64 synonymous
mutations per sample** in TCGA-LUAD (530 samples). Since the true `bmr`
(`sc = 359000`) already yields ≈ 63 syn/sample, **`g = 64.2 / 63 ≈ 1.02`** — the
silent matrix and the ERBB3 mutations share **one realistic LUAD-scale
background**, run on the **same simulated cohorts** as the count-based methods.

The estimated-background helpers are in
[`code/simulate_silent_estimate.R`](https://github.com/szhaolab/diffdriver-paper/blob/main/code/simulate_silent_estimate.R):

- **DiffDriver** estimates its background by running its signature model
  (`fastTopics` on the 96 × N silent count matrix, exactly as
  `diffdriver::matrixlistToBMR` does) to recover the per-sample per-context
  spectrum (`diffdriver_estimated_mr()`).
- **coselens** estimates its per-context substitution rate as counts / sites —
  `N_syn / Lsyn96` per group (`coselens_estimated_rate()`) — which is exactly the
  Poisson fit `dndscv` performs; this rate is fed to `dndscv_truerate`.

Per iteration:

```{r eval=FALSE}
source("code/simulate_functions.R"); source("code/simulate_silent_estimate.R")
source("code/run_coselens.R"); source("code/dndscv_truerate.R")
library(diffdriver); library(coselens)
load("data/bmr_sim_inputs.rda"); load("data/hmm.rda")
load("data/codeSignature.rda");  load("data/hotspot1sig.rda")
cm <- readRDS("data/context96_map.rds"); Lsyn96 <- readRDS("data/Lsyn.rds")$Lsyn96

sim <- simulate_confounding(binary = TRUE, sganno = bmr_sim$sganno,
  sgmatrix = bmr_sim$sgmatrix, Nsample = 1500, para = c(0.5, 0.5),
  beta_gc = bmr_sim$beta_gc, signatures = bmr_sim$signature, loadings = bmr_sim$ll,
  rho = sqrt(0.78), hot = 1, hmm = hmm, sc = 359000)
bmr <- sim$bmrfold$bmr; e <- sim$pheno; g1 <- which(e == 1); g2 <- which(e == 0)
maf <- sim_to_maf(sim$mutlist, sim$annodata, e)

Y <- build_silent_matrix(bmr, Lsyn96, target_silent = 64.2)   # ~64 syn/sample

# coselens: estimated per-group rate from the silent counts
mr1 <- coselens_estimated_rate(Y, g1, Lsyn96, cm)
mr2 <- coselens_estimated_rate(Y, g2, Lsyn96, cm)
cos <- run_coselens(maf$group1, maf$group2, subset.genes.by = "ERBB3",
         true_rate = list(mutrates1 = mr1, mutrates2 = mr2, theta = 1e6),
         gene_list = "ERBB3", cv = NULL)$substitutions$pval

# DiffDriver: estimated per-site background from fastTopics(Y)
mr_est <- do.call(rbind, diffdriver_estimated_mr(Y, sim$bmrmtxlist, bmr))
dd <- ddmodel_binary(do.call(rbind, sim$mutlist), e, mr_est, sim$efsize$diffFe)$pvalue
```

The full benchmark loops this over `Nsample = 200 … 1500` (100 iterations each)
and stores the estimated-background results in `data/pppower_estimatedBMR_bmr.Rd`
(the `coselens_est` and `diff_est` fields plotted above).
